### 2022-09-29 types.tex
infrastructure

\documentclass{article}
\usepackage{utf8math,ttquot,mathpartir,amsmath, amssymb, hydrocomments, mathtools,multicol}
\usepackage[margin=1in]{geometry}
\begin{document}

Writing things up in LaTeX, finally

\end{document}

### 2022-09-30 types.tex
wip
math blackboard N reinsertion

@@
 \documentclass{article}
-\usepackage{utf8math,ttquot,mathpartir,amsmath, amssymb, hydrocomments, mathtools,multicol}
+\usepackage{utf8math,ttquot,mathpartir,amsmath, amssymb, hydrocomments, mathtools,multicol,xspace}
 \usepackage[margin=1in]{geometry}
@@
 
-Writing things up in LaTeX, finally
+\section{syntax}
+
+\newcommand{\closedprogram}{\textit{closed-program}\xspace}
+\newcommand{\compiledcomponent}{\textit{compiled-component}\xspace}
+\newcommand{\incast}{\textit{incast}\xspace}
+\newcommand{\outcast}{\textit{outcast}\xspace}
+\newcommand{\seqstart}{\textit{seq-start}\xspace}
+\newcommand{\seqend}{\textit{seq-end}\xspace}
+\newcommand{\chain}{\textit{chain}\xspace}
+\newcommand{\op}{\textit{op}\xspace}
+\newcommand{\opt}{τ_{\textit{op}}\xspace}
+\newcommand{\N}{ℕ}
+\newcommand{\fresh}{\textit{fresh}\xspace}
+
+\begin{figure}
+  \begin{align*}
+    \closedprogram &:= "declare"~ A…~"in"~ \compiledcomponent…\\
+    A &∈ \textit{handoff-names}\\
+    \compiledcomponent &:= \incast;\outcast ∣ [\incast ∣ …]\op[\outcast ∣ …] \\
+    \incast &= A; \chain ∣ [\incast ∣ …]\op;\chain\\
+    \outcast &= \chain; A ∣ \chain; \op[\outcast ∣ …]\\
+  \end{align*}
+  \label{fig:syntax}
+  \caption{Syntax}
+\end{figure}
+
+In figure \ref{fig:syntax}
+
+\section{types}
+
+\begin{figure}
+    \begin{mathpar}
+      \inferrule{C,A : d ⊢ "declare"~\textit{rest}…~"in"~\compiledcomponent}{C ⊢ "declare"~A, \textit{rest}…~"in"~\compiledcomponent}
+
+      \inferrule{C ⊢ \incast \\ C ⊢ \outcast}{C ⊢ "declare"~"in"~\incast;\outcast}
+
+      \inferrule{C ⊢ [\incast ∣ …]\op_{α} \\ C ⊢ \op_{α}[\outcast ∣ …] \\ α~\fresh \\\text{something for operator}}{C ⊢ "declare"~"in"~[\incast…]\op[\outcast…]}
+
+    \end{mathpar}
+    \includegraphics{freehand-types}
+  \label{fig:types}
+  \caption{Types}
+\end{figure}
+
+I feel like we need to put operators in the environment if we want to
+be able to do the incast-outcast split. In fact, I feel like we might
+have a derivation problem _in general_ when doing the incast-outcast
+split. Annoying!
+
+In figure \ref{fig:types}
 

### 2022-10-11 types.tex
more progress writing things up

@@
 \newcommand{\fresh}{\textit{fresh}\xspace}
+\newcommand{\inputs}{\textit{inputs}\xspace}
+\newcommand{\outputs}{\textit{outputs}\xspace}
 
@@
   \begin{align*}
-    \closedprogram &:= "declare"~ A…~"in"~ \compiledcomponent…\\
+    \closedprogram &:= "declare"~ A…~"in"~ p\\
     A &∈ \textit{handoff-names}\\
-    \compiledcomponent &:= \incast;\outcast ∣ [\incast ∣ …]\op[\outcast ∣ …] \\
-    \incast &= A; \chain ∣ [\incast ∣ …]\op;\chain\\
-    \outcast &= \chain; A ∣ \chain; \op[\outcast ∣ …]\\
+    p &:= [p]\op[p] ∣ A ∣ (p|p)
   \end{align*}
@@
   \begin{mathpar}
-      \inferrule{C,A : d ⊢ "declare"~\textit{rest}…~"in"~\compiledcomponent}{C ⊢ "declare"~A, \textit{rest}…~"in"~\compiledcomponent}
 
-      \inferrule{C ⊢ \incast \\ C ⊢ \outcast}{C ⊢ "declare"~"in"~\incast;\outcast}
+    {\inferrule{A : in(dₐ : τₐ), A : out(dₐ | τₐ),… ;. : ⊤ ⊢ p : τ}{⊢ "declare"~A…~"in"~p : τ}}
 
-      \inferrule{C ⊢ [\incast ∣ …]\op_{α} \\ C ⊢ \op_{α}[\outcast ∣ …] \\ α~\fresh \\\text{something for operator}}{C ⊢ "declare"~"in"~[\incast…]\op[\outcast…]}
+    {
+      \inferrule{\inputs(\op) = τ' \\ \outputs(\op) = τ_o \\
+        C;. : ⊤ ⊢ p : τ' \\ C;d.p.\op[0] :^r τ_o ⊢ p' : τ_o'}
+                {{C,C'};{d : τ} ⊢ [p]\op[p'] : τ_o'}}
+
+    {\inferrule[relevant-parallel]{C₁;d[n] : τ ⊢ p₁ : τ₁ \\ C₂;d[n+1] :^r τ ⊢ p₂ : τ₂}{C₁,C₂;d[n] :^r τ ⊢ (p₁ ∣ p₂) : (τ₁ ∣ τ₂)}}
+
+    {\inferrule[substructural-fallback]{C;d : τ' ⊢ p : τ}{C;d : τ' ⊢^r p : τ}}
+
+    {\inferrule[general-parallel]{C₁;. : ⊤ ⊢ p₁ : τ₁ \\ C₂;. : ⊤ ⊢ p₂ : τ₂}{C₁,C₂; . : ⊤ ⊢ (p₁ ∣ p₂) : (τ₁ ∣ τ₂)}}
+
+    {\inferrule[exchange]{A : (dₐ : τₐ),  B : (d_b | τ_b) ; d : τ ⊢ p : τ}{B : (d_b | τ_b), A : (dₐ : τₐ) ⊢ p : τ}}
+
+    {\inferrule[unfold]{A:d',C;d[A/d] : τ ⊢ p}{A:d',C;d : τ ⊢ p}}
+
+    {\inferrule[fold]{A:d',C;d : τ ⊢ p}{A:d',C;d[A/d] : τ ⊢ p}}
+
+    {\inferrule[varref-out]{}{A:out(d : τ);d : τ ⊢ A : τ}}
+
+    {\inferrule[varref-in]{}{A:in(d : τ);. : \_ ⊢ A : τ}}
 
     \end{mathpar}
-    \includegraphics{freehand-types}
+    %\includegraphics{freehand-types}
   \label{fig:types}
@@
 
-I feel like we need to put operators in the environment if we want to
-be able to do the incast-outcast split. In fact, I feel like we might
-have a derivation problem _in general_ when doing the incast-outcast
-split. Annoying!
-
-In figure \ref{fig:types}
+We're not bothering with the oaklandish split right now, it's too silly.
+Open question if we need the relevance version of the rule, or can remove the $:^r$ annotation and observe that the $[n]$ version is only ever used with empty contexts for inputs to operators.
+Also worth wondering if we're going to be in trouble overloading the [] syntax here; time will tell.
+Types appear in figure \ref{fig:types}
 

### 2022-10-13 types.tex
wip

@@
 \begin{document}
+
+\abstract{Adding properties to hydroflow is important, because it
+  allows us to do three things: (1) it lets us explicitly state and
+  verify the requirements individual hydroflow operators have assumed
+  of their inputs in order to be correct, (2) It allows us to encode
+  translational goals and optimization correctness conditions, and (3) }
+
+\section{technical todo}
+
+\begin{itemize}
+\item Type system: basic structure and derivations ✓
+\item Type system: properties and their satisfaction
+\item semantics: core semantic model (is it rewrite-based?)
+\item Type system: axiom rewrite system
+\item core language: axiom extensibility
+\item proof: progress+preservation (or its equivalent in axiomatic semantics)
+\item proof: axiom admissibility (via bisimulation?)
+\item evaluation: example programs which satisfy the above
+\item evaluation: programmability assessment
+\item evaluation: extensibility assessment (new operators)
+\end{itemize}
 
@@
 
-    {\inferrule{A : in(dₐ : τₐ), A : out(dₐ | τₐ),… ;. : ⊤ ⊢ p : τ}{⊢ "declare"~A…~"in"~p : τ}}
+    {\inferrule{A : in(dₐ : τₐ), A : out(dₐ | τₐ),… ;. : ⊤ ⊢ p : τ}
+      {⊢ "declare"~A…~"in"~p : τ}}
 
@@
 
-    {\inferrule[relevant-parallel]{C₁;d[n] : τ ⊢ p₁ : τ₁ \\ C₂;d[n+1] :^r τ ⊢ p₂ : τ₂}{C₁,C₂;d[n] :^r τ ⊢ (p₁ ∣ p₂) : (τ₁ ∣ τ₂)}}
+    {\inferrule[relevant-parallel]
+      {C₁;d[n] : τ ⊢ p₁ : τ₁ \\ C₂;d[n+1] :^r τ ⊢ p₂ : τ₂}
+      {C₁,C₂;d[n] :^r τ ⊢ (p₁ ∣ p₂) : (τ₁ ∣ τ₂)}}
 
-    {\inferrule[substructural-fallback]{C;d : τ' ⊢ p : τ}{C;d : τ' ⊢^r p : τ}}
+    {\inferrule[substructural-fallback]
+      {C;d : τ' ⊢ p : τ}
+      {C;d : τ' ⊢^r p : τ}}
 
-    {\inferrule[general-parallel]{C₁;. : ⊤ ⊢ p₁ : τ₁ \\ C₂;. : ⊤ ⊢ p₂ : τ₂}{C₁,C₂; . : ⊤ ⊢ (p₁ ∣ p₂) : (τ₁ ∣ τ₂)}}
+    {\inferrule[general-parallel]
+      {C₁;. : ⊤ ⊢ p₁ : τ₁ \\ C₂;. : ⊤ ⊢ p₂ : τ₂}
+      {C₁,C₂; . : ⊤ ⊢ (p₁ ∣ p₂) : (τ₁ ∣ τ₂)}}
 
-    {\inferrule[exchange]{A : (dₐ : τₐ),  B : (d_b | τ_b) ; d : τ ⊢ p : τ}{B : (d_b | τ_b), A : (dₐ : τₐ) ⊢ p : τ}}
+    {\inferrule[exchange]
+      {A : (dₐ : τₐ),  B : (d_b | τ_b) ; d : τ ⊢ p : τ}
+      {B : (d_b | τ_b), A : (dₐ : τₐ) ⊢ p : τ}}
 
@@
 
-We're not bothering with the oaklandish split right now, it's too silly.
-Open question if we need the relevance version of the rule, or can remove the $:^r$ annotation and observe that the $[n]$ version is only ever used with empty contexts for inputs to operators.
-Also worth wondering if we're going to be in trouble overloading the [] syntax here; time will tell.
-Types appear in figure \ref{fig:types}
+We're not bothering with the oaklandish split right now, it's too
+silly.  Open question if we need the relevance version of the rule, or
+can remove the $:^r$ annotation and observe that the $[n]$ version is
+only ever used with empty contexts for inputs to operators. Also worth
+wondering if we're going to be in trouble overloading the [] syntax
+here; time will tell. Types appear in figure \ref{fig:types}
 

### 2022-10-13 types.tex
general notes on weakening

@@
 
+\section{general notes}
+
+Something we originally wanted to do was relaxations---operators that happened to consume input $'s$, but could just as well accept any $"permute"('s)$.
+How do we want to model this relaxation now, and is it still in scope?
+Concretely, I believe that this _is_ still in scope, and takes the form of setting translation-preservation targets to literal expressions like $"permute"('s)$.
+The challenge here will be reasoning in the algebra; we're not talking about equational rewriting anymore, since we also need to include weakening ("permute" generalizes lots of transformations).
+
 \end{document}

### 2022-10-18 types.tex
some typo correction
a major change: new context, removal of types from the old context, removal of types from the derivation term. Very unclear things are sound right now.

@@
 \newcommand{\outputs}{\textit{outputs}\xspace}
+\newcommand{\IT}[1]{{\textit{#1}}\xspace}
 
@@
     A &∈ \textit{handoff-names}\\
-    p &:= [p]\op[p] ∣ A ∣ (p|p)
+    p_\IT{out} &:= [p_\IT{out}]\op ∣ A ∣ (p_\IT{out}|p_\IT{out})\\
+    p_\IT{in} &:= [p_\IT{in}]\op ∣ A ∣ (p_\IT{in}|p_\IT{in})\\
+    p &:= [p_\IT{in}]\op[p_\IT{out}] ∣ (p|p)
   \end{align*}
@@
 
-    {\inferrule{A : in(dₐ : τₐ), A : out(dₐ | τₐ),… ;. : ⊤ ⊢ p : τ}
+    {\inferrule{A : in(dₐ), A : out(dₐ),… ;A : τₐ,…;.  ⊢ p : τ}
       {⊢ "declare"~A…~"in"~p : τ}}
@@
     {
-      \inferrule{\inputs(\op) = τ' \\ \outputs(\op) = τ_o \\
-        C;. : ⊤ ⊢ p : τ' \\ C;d.p.\op[0] :^r τ_o ⊢ p' : τ_o'}
-                {{C,C'};{d : τ} ⊢ [p]\op[p'] : τ_o'}}
+      \inferrule{\inputs(\op) = τ \\ 
+        C;Γ;. ⊢ p : τ \\ C';Γ';d.p.\op[0] ⊢ p' : τ'}
+                {{C,C'};Γ,Γ';{d} ⊢ [p]\op[p'] : τ'}}
 
     {\inferrule[relevant-parallel]
-      {C₁;d[n] : τ ⊢ p₁ : τ₁ \\ C₂;d[n+1] :^r τ ⊢ p₂ : τ₂}
-      {C₁,C₂;d[n] :^r τ ⊢ (p₁ ∣ p₂) : (τ₁ ∣ τ₂)}}
-
-    {\inferrule[substructural-fallback]
-      {C;d : τ' ⊢ p : τ}
-      {C;d : τ' ⊢^r p : τ}}
+      {C₁;Γ_1;d[n] ⊢ p₁ : τ₁ \\ C₂;Γ_2;d[n+1] ⊢ p₂ : τ₂}
+      {C₁,C₂;Γ_1,Γ_2;d[n] ⊢ (p₁ ∣ p₂) : (τ₁ ∣ τ₂)}}
 
     {\inferrule[general-parallel]
-      {C₁;. : ⊤ ⊢ p₁ : τ₁ \\ C₂;. : ⊤ ⊢ p₂ : τ₂}
-      {C₁,C₂; . : ⊤ ⊢ (p₁ ∣ p₂) : (τ₁ ∣ τ₂)}}
+      {C₁;Γ_1;. ⊢ p₁ : τ₁ \\ C₂;Γ_2;. ⊢ p₂ : τ₂}
+      {C₁,C₂;Γ_1,Γ_2; . ⊢ (p₁ ∣ p₂) : (τ₁ ∣ τ₂)}}
 
-    {\inferrule[exchange]
-      {A : (dₐ : τₐ),  B : (d_b | τ_b) ; d : τ ⊢ p : τ}
-      {B : (d_b | τ_b), A : (dₐ : τₐ) ⊢ p : τ}}
+    {\inferrule[exchange-C]
+      {C,C_2,C_1,C' ;Γ; d ⊢ p : τ}
+      {C,C_1,C_2,C' ;Γ; d ⊢ p : τ}}
 
-    {\inferrule[unfold]{A:d',C;d[A/d] : τ ⊢ p}{A:d',C;d : τ ⊢ p}}
+    {\inferrule[exchange-gamma]
+      {C ;Γ,Γ_1,Γ_2,Γ' ; d ⊢ p : τ}
+      {C ;Γ,Γ_2,Γ_1,Γ'; d ⊢ p : τ}}
 
-    {\inferrule[fold]{A:d',C;d : τ ⊢ p}{A:d',C;d[A/d] : τ ⊢ p}}
+    {\inferrule[weaken-gamma]
+      {C ;Γ,Γ' ; d ⊢ p : τ}
+      {C ;Γ,Γ_w,Γ'; d ⊢ p : τ}}
 
-    {\inferrule[varref-out]{}{A:out(d : τ);d : τ ⊢ A : τ}}
+    {\inferrule[contract-gamma]
+      {C ;Γ,Γ_c,Γ_c,Γ' ; d ⊢ p : τ}
+      {C ;Γ,Γ_c,Γ'; d ⊢ p : τ}}
 
-    {\inferrule[varref-in]{}{A:in(d : τ);. : \_ ⊢ A : τ}}
+    {\inferrule[unfold]{A:d',C;Γ;d[A/d] ⊢ p : τ}{A:d',C;Γ;d ⊢ p : τ}}
 
-    \end{mathpar}
+    {\inferrule[fold]{A:d',C;Γ;d⊢ p : τ}{A:d',C;Γ;d[A/d]⊢ p : τ}}
+
+    {\inferrule[varref-out]{Γ ⊢ d : τ}{A:out(d);Γ;d ⊢ A : τ}}
+
+    {\inferrule[varref-in]{}{A:in(d);A : τ; . ⊢ A : τ}}
+
+    \end{mathpar}\MPM{There's something fishy going on with varref-in; right now, you could use it and then immediately apply unfold to get a derivation which might not typecheck.
+We need to prove this is not actually a problem.}
     %\includegraphics{freehand-types}

### 2022-10-19 types.tex
more updates; things look dicey tbh

@@
 
+    {\inferrule[axoim-transform]{Γ ⊢ d ⇝ d' \\ C;Γ;d' ⊢ p : τ}{C;Γ;d ⊢ p : τ}}
+
     \end{mathpar}\MPM{There's something fishy going on with varref-in; right now, you could use it and then immediately apply unfold to get a derivation which might not typecheck.
@@
   \label{fig:types}
-  \caption{Types}
+  \caption{Types: top-level}
+\end{figure}
+\begin{figure}
+\begin{mathpar}
+
+    {\inferrule[derivation-type-base]{}{Γ ⊢ d : τ}}
+
+\end{mathpar}    
+    \caption{Types: derivations }
+    \label{fig:types2}
+\end{figure}
+\begin{figure}
+\begin{mathpar}
+
+{\inferrule[axoim-transform-base]{}{Γ ⊢ d ⇝ d'}}
+
+\end{mathpar}    
+    \caption{Types: axiom transformations }
+    \label{fig:types3}
 \end{figure}
 
-We're not bothering with the oaklandish split right now, it's too
-silly.  Open question if we need the relevance version of the rule, or
-can remove the $:^r$ annotation and observe that the $[n]$ version is
-only ever used with empty contexts for inputs to operators. Also worth
-wondering if we're going to be in trouble overloading the [] syntax
-here; time will tell. Types appear in figure \ref{fig:types}
+Worth wondering if we're going to be in trouble overloading the $[]$ syntax here; time will tell. 
+Types appear in figure \ref{fig:types}. 
+Another concern is that we have a uhhh _lot_ of type systems in the mix at this point: 
+we've got $C;Γ;d ⊢ p : τ$, $Γ ⊢ d : τ$, and $Γ ⊢ d ⇝ d'$.  
+That's a lot more independent systems than we'd ideally  have.  Is there a way to collapse this back down?
 

### 2022-10-20 types.tex
some late-night realization notes

@@
 
+Another important note!
+I think fold/unfold are wrong and unfit for purpose, actually.
+We should really be replacing the variable names with their derivation expression when they're placed into the derivation context for the first time.
+But as of right now, that's not how the type system works.
+Maybe we need to "output" a derivation, rather than put it in the context.
+Note: doing this naively immediately breaks on the dispatch to relevant-parallel and general-parallel.
+I think it still holds significant promise, but might not be viable for 11:40pm.
+
 \section{general notes}

### 2022-10-20 types.tex
removing fold/unfold points to a need to treat ingress/egress differently. This in turn makes the overall type system duplicate a bunch of rules but also make more sense.
If we want to just have one, it looks like "just have egress" is the one that hangs together the most right now.

@@
 
-    {\inferrule{A : in(dₐ), A : out(dₐ),… ;A : τₐ,…;.  ⊢ p : τ}
-      {⊢ "declare"~A…~"in"~p : τ}}
+    {\inferrule{A : in(dₐ), A : out(dₐ),… ;A : τₐ,…  ⊢ p}
+      {⊢ "declare"~A…~"in"~p}}
 
     {
-      \inferrule{\inputs(\op) = τ \\ 
-        C;Γ;. ⊢ p : τ \\ C';Γ';d.p.\op[0] ⊢ p' : τ'}
-                {{C,C'};Γ,Γ';{d} ⊢ [p]\op[p'] : τ'}}
+      \inferrule[pivot]{\inputs(\op) = τ \\ 
+        C;Γ ⊢ p : d~τ \\ C';Γ';d.op[0] ⊢ p'}
+                {{C,C'};Γ,Γ' ⊢ [p]\op[p']}}
 
-    {\inferrule[relevant-parallel]
-      {C₁;Γ_1;d[n] ⊢ p₁ : τ₁ \\ C₂;Γ_2;d[n+1] ⊢ p₂ : τ₂}
-      {C₁,C₂;Γ_1,Γ_2;d[n] ⊢ (p₁ ∣ p₂) : (τ₁ ∣ τ₂)}}
+    {
+      \inferrule[inward]{\inputs(\op) = τ \\ 
+        C;Γ ⊢ p : d~τ}
+                {{C};Γ,Γ' ⊢ [p]\op : d.\op~τ'}}
 
-    {\inferrule[general-parallel]
-      {C₁;Γ_1;. ⊢ p₁ : τ₁ \\ C₂;Γ_2;. ⊢ p₂ : τ₂}
-      {C₁,C₂;Γ_1,Γ_2; . ⊢ (p₁ ∣ p₂) : (τ₁ ∣ τ₂)}}
+{
+      \inferrule[outward]{Γ ⊢ d : τ \\ \inputs(\op) = τ \\
+        C;Γ;d.\op[0] ⊢ p'}
+                {{C};Γ;d ⊢ \op[p']}}
 
-    {\inferrule[exchange-C]
-      {C,C_2,C_1,C' ;Γ; d ⊢ p : τ}
-      {C,C_1,C_2,C' ;Γ; d ⊢ p : τ}}
+    {\inferrule[out-parallel]
+      {C₁;Γ_1;d[n] ⊢ p₁ \\ C₂;Γ_2;d[n+1] ⊢ p₂}
+      {C₁,C₂;Γ_1,Γ_2;d[n] ⊢ (p₁ ∣ p₂)}}
 
-    {\inferrule[exchange-gamma]
-      {C ;Γ,Γ_1,Γ_2,Γ' ; d ⊢ p : τ}
-      {C ;Γ,Γ_2,Γ_1,Γ'; d ⊢ p : τ}}
+    {\inferrule[in-parallel]
+      {C₁;Γ_1 ⊢ p₁ : d₁~τ₁ \\ C₂;Γ_2 ⊢ p₂ : d₂~τ₂}
+      {C₁,C₂;Γ_1,Γ_2 ⊢ (p₁ ∣ p₂) : (d₁ ∣ d₂)~(τ₁ ∣ τ₂)}}
 
-    {\inferrule[weaken-gamma]
-      {C ;Γ,Γ' ; d ⊢ p : τ}
-      {C ;Γ,Γ_w,Γ'; d ⊢ p : τ}}
+%    {\inferrule[unfold]{A:d',C;Γ;d[A/d] ⊢ p : τ}{A:d',C;Γ;d ⊢ p : τ}}
 
-    {\inferrule[contract-gamma]
-      {C ;Γ,Γ_c,Γ_c,Γ' ; d ⊢ p : τ}
-      {C ;Γ,Γ_c,Γ'; d ⊢ p : τ}}
+%    {\inferrule[fold]{A:d',C;Γ;d⊢ p : τ}{A:d',C;Γ;d[A/d]⊢ p : τ}}
 
-    {\inferrule[unfold]{A:d',C;Γ;d[A/d] ⊢ p : τ}{A:d',C;Γ;d ⊢ p : τ}}
+    {\inferrule[varref-out]{Γ ⊢ d : τ}{A:out(d);Γ;d ⊢ A}}
 
-    {\inferrule[fold]{A:d',C;Γ;d⊢ p : τ}{A:d',C;Γ;d[A/d]⊢ p : τ}}
+    {\inferrule[varref-in]{}{A:in(d);A : τ ⊢ A : d~τ}}
 
-    {\inferrule[varref-out]{Γ ⊢ d : τ}{A:out(d);Γ;d ⊢ A : τ}}
+    {\inferrule[axoim-transform-in]{Γ ⊢ d ⇝ d' \\ C;Γ ⊢ p : d~τ}{C;Γ ⊢ p : d'~τ}}
 
-    {\inferrule[varref-in]{}{A:in(d);A : τ; . ⊢ A : τ}}
-
-    {\inferrule[axoim-transform]{Γ ⊢ d ⇝ d' \\ C;Γ;d' ⊢ p : τ}{C;Γ;d ⊢ p : τ}}
+    {\inferrule[axoim-transform-out]{Γ ⊢ d ⇝ d' \\ C;Γ;d' ⊢ p}{C;Γ;d ⊢ p}}
 
@@
 \end{figure}
+
 \begin{figure}
@@
     \label{fig:types3}
+\end{figure}
+
+\begin{figure}
+\begin{mathpar}
+
+    {\inferrule[exchange-C-in]
+      {C,C_2,C_1,C' ;Γ ⊢ p : d~τ}
+      {C,C_1,C_2,C' ;Γ ⊢ p : d~τ}}
+
+    {\inferrule[exchange-gamma-in]
+      {C ;Γ,Γ_1,Γ_2,Γ' ⊢ p : d~τ}
+      {C ;Γ,Γ_2,Γ_1,Γ' ⊢ p : d~τ}}
+
+    {\inferrule[weaken-gamma-in]
+      {C ;Γ,Γ'  ⊢ p : d~τ}
+      {C ;Γ,Γ_w,Γ' ⊢ p : d~τ}}
+
+    {\inferrule[contract-gamma-in]
+      {C ;Γ,Γ_c,Γ_c,Γ' ⊢ p : d~τ}
+      {C ;Γ,Γ_c,Γ' ⊢ p : d~τ}}
+
+    {\inferrule[exchange-C-out]
+      {C,C_2,C_1,C' ;Γ; d ⊢ p}
+      {C,C_1,C_2,C' ;Γ; d ⊢ p}}
+
+    {\inferrule[exchange-gamma-out]
+      {C ;Γ,Γ_1,Γ_2,Γ'; d ⊢ p}
+      {C ;Γ,Γ_2,Γ_1,Γ'; d ⊢ p}}
+
+    {\inferrule[weaken-gamma-out]
+      {C ;Γ,Γ'; d  ⊢ p}
+      {C ;Γ,Γ_w,Γ'; d ⊢ p}}
+
+    {\inferrule[contract-gamma-out]
+      {C ;Γ,Γ_c,Γ_c,Γ'; d ⊢ p}
+      {C ;Γ,Γ_c,Γ'; d ⊢ p}}
+
+\end{mathpar}    
+    \caption{Types: boring structural stuff }
+    \label{fig:types-structural}
 \end{figure}
@@
 
+Another note: the "natural" way to do derivations might change between a push-oriented and a pull-oriented worldview---worth thinking about / playing around with.
+
+On the above: turns out I did stay up and push it through and... it seems to work! Shockingly! Quite Interesting.
+
 \section{general notes}

### 2022-10-24 types.tex
another system overhaul, maybe with its own bugs.

--- 2022-10-20 types.tex
@@
 
-    {\inferrule{A : in(dₐ), A : out(dₐ),… ;A : τₐ,…;.  ⊢ p : τ}
-      {⊢ "declare"~A…~"in"~p : τ}}
+    {\inferrule{A : in(dₐ'.dₐ''), A : out(dₐ.dₐ''),… ⊢ p
+    \\ dₐ' ∈ \{A.dₐ'', dₐ.dₐ''\}
+    \\ (A : τₐ,… ⊢ dₐ.dₐ'' : τₐ)…
+    }
+      {⊢ "declare"~A…~"in"~p}}
 
     {
-      \inferrule{\inputs(\op) = τ \\ 
-        C;Γ;. ⊢ p : τ \\ C';Γ';d.p.\op[0] ⊢ p' : τ'}
-                {{C,C'};Γ,Γ';{d} ⊢ [p]\op[p'] : τ'}}
+      \inferrule[pivot]{|\outputs(\op)| = |p'| \\ 
+        C ⊢ p : d \\ C';d.op[0] ⊢ p'}
+                {{C,C'} ⊢ [p]\op[p']}}
 
-    {\inferrule[relevant-parallel]
-      {C₁;Γ_1;d[n] ⊢ p₁ : τ₁ \\ C₂;Γ_2;d[n+1] ⊢ p₂ : τ₂}
-      {C₁,C₂;Γ_1,Γ_2;d[n] ⊢ (p₁ ∣ p₂) : (τ₁ ∣ τ₂)}}
+    {
+      \inferrule[inward]{C ⊢ p : d}
+                {{C} ⊢ [p]\op : d.\op[0]}}
 
-    {\inferrule[general-parallel]
-      {C₁;Γ_1;. ⊢ p₁ : τ₁ \\ C₂;Γ_2;. ⊢ p₂ : τ₂}
-      {C₁,C₂;Γ_1,Γ_2; . ⊢ (p₁ ∣ p₂) : (τ₁ ∣ τ₂)}}
+{
+      \inferrule[outward]{|\outputs(\op)| = |p'| \\
+        C;d.\op[0] ⊢ p'}
+                {{C};d ⊢ \op[p']}}
 
-    {\inferrule[exchange-C]
-      {C,C_2,C_1,C' ;Γ; d ⊢ p : τ}
-      {C,C_1,C_2,C' ;Γ; d ⊢ p : τ}}
+    {\inferrule[out-parallel]
+      {C₁;d[n] ⊢ p₁ \\ C₂;d[n+1] ⊢ p₂}
+      {C₁,C₂;d[n] ⊢ (p₁ ∣ p₂)}}
 
-    {\inferrule[exchange-gamma]
-      {C ;Γ,Γ_1,Γ_2,Γ' ; d ⊢ p : τ}
-      {C ;Γ,Γ_2,Γ_1,Γ'; d ⊢ p : τ}}
+    {\inferrule[in-parallel]
+      {C₁ ⊢ p₁ : d₁ \\ C₂ ⊢ p₂ : d₂}
+      {C₁,C₂ ⊢ (p₁ ∣ p₂) : (d₁ ∣ d₂)}}
 
-    {\inferrule[weaken-gamma]
-      {C ;Γ,Γ' ; d ⊢ p : τ}
-      {C ;Γ,Γ_w,Γ'; d ⊢ p : τ}}
+    {\inferrule[varref-out]{}{A:out(d);d ⊢ A}}
 
-    {\inferrule[contract-gamma]
-      {C ;Γ,Γ_c,Γ_c,Γ' ; d ⊢ p : τ}
-      {C ;Γ,Γ_c,Γ'; d ⊢ p : τ}}
+    {\inferrule[varref-out-approx]{}{A:out(d'.d);B.d ⊢ A}}\textit{\MPM{Is this sound? Or can it allow the paths to vary too much? I feel like probably the latter....}}
 
-    {\inferrule[unfold]{A:d',C;Γ;d[A/d] ⊢ p : τ}{A:d',C;Γ;d ⊢ p : τ}}
+    {\inferrule[varref-in]{}{A:in(d) ⊢ A : d}}
 
-    {\inferrule[fold]{A:d',C;Γ;d⊢ p : τ}{A:d',C;Γ;d[A/d]⊢ p : τ}}
-
-    {\inferrule[varref-out]{Γ ⊢ d : τ}{A:out(d);Γ;d ⊢ A : τ}}
-
-    {\inferrule[varref-in]{}{A:in(d);A : τ; . ⊢ A : τ}}
-
-    {\inferrule[axoim-transform]{Γ ⊢ d ⇝ d' \\ C;Γ;d' ⊢ p : τ}{C;Γ;d ⊢ p : τ}}
-
-    \end{mathpar}\MPM{There's something fishy going on with varref-in; right now, you could use it and then immediately apply unfold to get a derivation which might not typecheck.
-We need to prove this is not actually a problem.}
-    %\includegraphics{freehand-types}
+    \end{mathpar}
   \label{fig:types}
@@
 \begin{mathpar}
+    {\inferrule[axoim-transform]{Γ ⊢ d ⇝ d' \\ Γ ⊢ d' : τ}{Γ ⊢ d : τ}}
 
@@
     \label{fig:types3}
+\end{figure}
+
+\begin{figure}
+\begin{mathpar}
+
+    {\inferrule[exchange-C-in]
+      {C,C_2,C_1,C' ;Γ ⊢ p : d}
+      {C,C_1,C_2,C' ;Γ ⊢ p : d}}
+
+    {\inferrule[exchange-C-out]
+      {C,C_2,C_1,C' ;Γ; d ⊢ p}
+      {C,C_1,C_2,C' ;Γ; d ⊢ p}}
+
+    {\inferrule[exchange-gamma]
+      {Γ,Γ_1,Γ_2,Γ' ⊢ φ}
+      {Γ,Γ_2,Γ_1,Γ' ⊢ φ}}
+
+    {\inferrule[weaken-gamma]
+      {Γ,Γ'  ⊢ φ}
+      {Γ,Γ_w,Γ' ⊢ φ}}
+
+    {\inferrule[contract-gamma]
+      {Γ,Γ_c,Γ_c,Γ' ⊢ φ}
+      {Γ,Γ_c,Γ' ⊢ φ}}
+
+\end{mathpar}    
+    \caption{Types: boring structural stuff }
+    \label{fig:types-structural}
 \end{figure}

### 2022-10-25 types.tex
some cleanup

@@
 Another concern is that we have a uhhh _lot_ of type systems in the mix at this point: 
-we've got $C;Γ;d ⊢ p : τ$, $Γ ⊢ d : τ$, and $Γ ⊢ d ⇝ d'$.  
+we've got $C;d ⊢ p$, $C ⊢ p : d$, $Γ ⊢ d : τ$, and $Γ ⊢ d ⇝ d'$.  
 That's a lot more independent systems than we'd ideally  have.  Is there a way to collapse this back down?
-
-Another important note!
-I think fold/unfold are wrong and unfit for purpose, actually.
-We should really be replacing the variable names with their derivation expression when they're placed into the derivation context for the first time.
-But as of right now, that's not how the type system works.
-Maybe we need to "output" a derivation, rather than put it in the context.
-Note: doing this naively immediately breaks on the dispatch to relevant-parallel and general-parallel.
-I think it still holds significant promise, but might not be viable for 11:40pm.
-
-Another note: the "natural" way to do derivations might change between a push-oriented and a pull-oriented worldview---worth thinking about / playing around with.
-
-On the above: turns out I did stay up and push it through and... it seems to work! Shockingly! Quite Interesting.
 

### 2022-10-26 types.tex
Updates from Overleaf

--- 2022-10-24 types.tex
@@
 \documentclass{article}
-\usepackage{utf8math,ttquot,mathpartir,amsmath, amssymb, hydrocomments, mathtools,multicol,xspace}
+\usepackage{utf8math,ttquot,mathpartir,amsmath, amssymb, hydrocomments, mathtools,multicol,xspace,stmaryrd }
 \usepackage[margin=1in]{geometry}
@@
 \newcommand{\outputs}{\textit{outputs}\xspace}
+\newcommand{\outarity}{\textit{output-arity}\xspace}
 \newcommand{\IT}[1]{{\textit{#1}}\xspace}
@@
 
-    {\inferrule{A : in(dₐ'.dₐ''), A : out(dₐ.dₐ''),… ⊢ p
-    \\ dₐ' ∈ \{A.dₐ'', dₐ.dₐ''\}
-    \\ (A : τₐ,… ⊢ dₐ.dₐ'' : τₐ)…
+    {\inferrule{A : in(dₐ' ▷ dₐ''), A : out(dₐ ▷ dₐ''),… ⊢ p
+    \\ dₐ' ∈ \{A ▷ dₐ'', dₐ ▷ dₐ''\}
+    \\ (A : τₐ,… ⊢ dₐ ▷ dₐ'' : τₐ)…
     }
@@
       \inferrule[pivot]{|\outputs(\op)| = |p'| \\ 
-        C ⊢ p : d \\ C';d.op[0] ⊢ p'}
+        C ⊢ p : d \\ C';d ▷ op[0] ⊢ p'}
                 {{C,C'} ⊢ [p]\op[p']}}
@@
       \inferrule[inward]{C ⊢ p : d}
-                {{C} ⊢ [p]\op : d.\op[0]}}
+                {{C} ⊢ [p]\op : d ▷ \op[0]}}
 
 {
-      \inferrule[outward]{|\outputs(\op)| = |p'| \\
-        C;d.\op[0] ⊢ p'}
+      \inferrule[outward]{\outarity(\op) = |p'| \\
+        C;d ▷ \op[0] ⊢ p'}
                 {{C};d ⊢ \op[p']}}
@@
 \begin{mathpar}
+
     {\inferrule[axoim-transform]{Γ ⊢ d ⇝ d' \\ Γ ⊢ d' : τ}{Γ ⊢ d : τ}}
 
-    {\inferrule[derivation-type-base]{}{Γ ⊢ d : τ}}
+    {\inferrule[parallel]
+      {Γ₁ ⊢ p₁ : τ₁ \\ Γ₂ ⊢ p₂ : τ₂}
+      {Γ₁,Γ₂ ⊢ (p₁ ∣ p₂) : (τ₁ ∣ τ₂)}}
+
+    {\inferrule[operator-typing]{Γ ⊢ d : τ'}{Γ ⊢ d ▷ \op[n] : (〚\op〛 τ')[n]}}
 
@@
     \caption{Types: derivations }
+    \label{fig:types2}
+\end{figure}
+
+\begin{figure}
+\begin{mathpar}
+
+\inferrule[action-on-type]{}{〚\op〛 τ ≜ τ'}
+
+\inferrule[index-into-type]{}{(τ₁ | … | τ_n | …)[n] ≜ τ_n}
+
+\end{mathpar}
+    \caption{Operators: actions on types }
     \label{fig:types2}

### 2022-10-26 types.tex
things are hanging together ok, but it seems likely they're not possible to do in Rust? Unclear.

@@
 
-    {\inferrule{A : in(dₐ' ▷ dₐ''), A : out(dₐ ▷ dₐ''),… ⊢ p
-    \\ dₐ' ∈ \{A ▷ dₐ'', dₐ ▷ dₐ''\}
-    \\ (A : τₐ,… ⊢ dₐ ▷ dₐ'' : τₐ)…
+    {\inferrule{A▷, dₐ▷A,… ⊢ p
+    \\ (dₐ▷A : τₐ,… ⊢ dₐ : τₐ)…
     }
@@
 
-    {\inferrule[varref-out]{}{A:out(d);d ⊢ A}}
+    {\inferrule[varref-out]{}{ d▷A;d ⊢ A}}
 
-    {\inferrule[varref-out-approx]{}{A:out(d'.d);B.d ⊢ A}}\textit{\MPM{Is this sound? Or can it allow the paths to vary too much? I feel like probably the latter....}}
+%    {\inferrule[varref-out-approx]{}{A:out(d'.d);B.d ⊢ A}}\textit{\MPM{Is this sound? Or can it allow the paths to vary too much? I feel like probably the latter....}}
 
-    {\inferrule[varref-in]{}{A:in(d) ⊢ A : d}}
+    {\inferrule[varref-in]{}{A▷ ⊢ A : A}}
 
@@
 
-    {\inferrule[axoim-transform]{Γ ⊢ d ⇝ d' \\ Γ ⊢ d' : τ}{Γ ⊢ d : τ}}
+    {\inferrule[assert-matching]{Γ ⊢ d ⇝ d' \\ Γ ⊢ d : τ}{Γ ⊢ d▷as(d')[0] : τ}}
 
@@
     {\inferrule[operator-typing]{Γ ⊢ d : τ'}{Γ ⊢ d ▷ \op[n] : (〚\op〛 τ')[n]}}
+
+    {\inferrule[varref]{}{ d▷A : τ ⊢ A : τ}}
 
@@
 
-{\inferrule[axoim-transform-base]{}{Γ ⊢ d ⇝ d'}}
+{\inferrule[axoim-transform-abstract]{Γ,dₐ▷A : τₐ ⊢ d ⇝ d' \\ A~\fresh}{Γ ⊢ d[A/dₐ] ⇝ d'}}
+
+{\inferrule[axoim-transform-concretize]{τₐ <: τ}{Γ,dₐ▷A : τₐ ⊢ d ⇝ d["?s"~τ/dₐ]}}
+
+{\inferrule[axoim-transform-transitive]{Γ ⊢ d ⇝ d' \\ Γ ⊢ d' ⇝ d''}{
+Γ ⊢ d ⇝ d''}}
+
+{\inferrule[axoim-transform-unfold]{}{Γ,dₐ▷A : τₐ  ⊢ d ⇝ d[A/dₐ]}}
+
+{\inferrule[axoim-transform-fold]{}{Γ,dₐ▷A : τₐ  ⊢ d[A/dₐ] ⇝ d}}
+
+{\inferrule[axoim-transform-reflexive]{}{Γ ⊢ d ⇝ d}}
 
@@
     \caption{Types: axiom transformations }
-    \label{fig:types3}
+    \label{fig:axiom-types}
 \end{figure}
@@
 
+A new concern: we're building out the axiom transformation system in figure \ref{fig:axiom-types}, which relies on these substitution operators.
+I'm really not sure how to encode the substitutions---or the notion of an environment into which substitutions occur at all---into Rust.
+I guess they'd be...
+literal variables?
+Very unclear.
+
 \section{general notes}

### 2022-10-29 paper.tex
maybe pldi format

\input{pldi23-preamble}

\begin{document}

\title{HydroFlow: A New Dataflow Programming Language}

\input{authors}

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
  A clear and well-documented \LaTeX\ document is presented as an
  article formatted for publication by ACM in a conference proceedings
  or journal publication. Based on the ``acmart'' document class, this
  article presents and explains many of the common variations, as well
  as many of the formatting elements an author may use in the
  preparation of the documentation of their work.
\end{abstract}

%% A "teaser" image appears between the author and affiliation
%% information and the body of the document, and typically spans the
%% page.
%\begin{teaserfigure}
%  \includegraphics[width=\textwidth]{sampleteaser}
%  \caption{Seattle Mariners at Spring Training, 2010.}
%  \Description{Enjoying the baseball game from the third-base
%  seats. Ichiro Suzuki preparing to bat.}
%  \label{fig:teaser}
%\end{teaserfigure}

%\received{20 February 2007}
%\received[revised]{12 March 2009}
%\received[accepted]{5 June 2009}

%%
%% This command processes the author and affiliation and title
%% information and builds the first part of the formatted document.
\maketitle

\section{Introduction}
ACM's consolidated article template, introduced in 2017, provides a
consistent \LaTeX\ style for use across ACM publications, and
incorporates accessibility and metadata-extraction functionality
necessary for future Digital Library endeavors. Numerous ACM and
SIG-specific \LaTeX\ templates have been examined, and their unique
features incorporated into this single new template.

If you are new to publishing with ACM, this document is a valuable
guide to the process of preparing your work for publication. If you
have published with ACM before, this document provides insight and
instruction into more recent changes to the article template.

The ``\verb|acmart|'' document class can be used to prepare articles
for any ACM publication --- conference or journal, and for any stage
of publication, from review to final ``camera-ready'' copy, to the
author's own version, with {\itshape very} few changes to the source.

\section{Template Overview}
As noted in the introduction, the ``\verb|acmart|'' document class can
be used to prepare many different kinds of documentation --- a
double-blind initial submission of a full-length technical paper, a
two-page SIGGRAPH Emerging Technologies abstract, a ``camera-ready''
journal article, a SIGCHI Extended Abstract, and more --- all by
selecting the appropriate {\itshape template style} and {\itshape
  template parameters}.

This document will explain the major features of the document
class. For further information, the {\itshape \LaTeX\ User's Guide} is
available from
\url{https://www.acm.org/publications/proceedings-template}.

\subsection{Template Styles}

The primary parameter given to the ``\verb|acmart|'' document class is
the {\itshape template style} which corresponds to the kind of publication
or SIG publishing the work. This parameter is enclosed in square
brackets and is a part of the {\verb|documentclass|} command:
\begin{verbatim}
  \documentclass[STYLE]{acmart}
\end{verbatim}

Journals use one of three template styles. All but three ACM journals
use the {\verb|acmsmall|} template style:
\begin{itemize}
\item {\texttt{acmsmall}}: The default journal template style.
\item {\texttt{acmlarge}}: Used by JOCCH and TAP.
\item {\texttt{acmtog}}: Used by TOG.
\end{itemize}

The majority of conference proceedings documentation will use the {\verb|acmconf|} template style.
\begin{itemize}
\item {\texttt{acmconf}}: The default proceedings template style.
\item{\texttt{sigchi}}: Used for SIGCHI conference articles.
\item{\texttt{sigplan}}: Used for SIGPLAN conference articles.
\end{itemize}

\subsection{Template Parameters}

In addition to specifying the {\itshape template style} to be used in
formatting your work, there are a number of {\itshape template parameters}
which modify some part of the applied template style. A complete list
of these parameters can be found in the {\itshape \LaTeX\ User's Guide.}

Frequently-used parameters, or combinations of parameters, include:
\begin{itemize}
\item {\texttt{anonymous,review}}: Suitable for a ``double-blind''
  conference submission. Anonymizes the work and includes line
  numbers. Use with the \texttt{\acmSubmissionID} command to print the
  submission's unique ID on each page of the work.
\item{\texttt{authorversion}}: Produces a version of the work suitable
  for posting by the author.
\item{\texttt{screen}}: Produces colored hyperlinks.
\end{itemize}

This document uses the following string as the first command in the
source file:
\begin{verbatim}
\documentclass[acmsmall]{acmart}
\end{verbatim}

\section{Modifications}

Modifying the template --- including but not limited to: adjusting
margins, typeface sizes, line spacing, paragraph and list definitions,
and the use of the \verb|\vspace| command to manually adjust the
vertical spacing between elements of your work --- is not allowed.

{\bfseries Your document will be returned to you for revision if
  modifications are discovered.}

\section{Typefaces}

The ``\verb|acmart|'' document class requires the use of the
``Libertine'' typeface family. Your \TeX\ installation should include
this set of packages. Please do not substitute other typefaces. The
``\verb|lmodern|'' and ``\verb|ltimes|'' packages should not be used,
as they will override the built-in typeface families.

\section{Title Information}

The title of your work should use capital letters appropriately -
\url{https://capitalizemytitle.com/} has useful rules for
capitalization. Use the {\verb|title|} command to define the title of
your work. If your work has a subtitle, define it with the
{\verb|subtitle|} command.  Do not insert line breaks in your title.

If your title is lengthy, you must define a short version to be used
in the page headers, to prevent overlapping text. The \verb|title|
command has a ``short title'' parameter:
\begin{verbatim}
  \title[short title]{full title}
\end{verbatim}

\section{Authors and Affiliations}

Each author must be defined separately for accurate metadata
identification.  As an exception, multiple authors may share one
affiliation. Authors' names should not be abbreviated; use full first
names wherever possible. Include authors' e-mail addresses whenever
possible.

Grouping authors' names or e-mail addresses, or providing an ``e-mail
alias,'' as shown below, is not acceptable:
\begin{verbatim}
  \author{Brooke Aster, David Mehldau}
  \email{dave,judy,steve@university.edu}
  \email{firstname.lastname@phillips.org}
\end{verbatim}

The \verb|authornote| and \verb|authornotemark| commands allow a note
to apply to multiple authors --- for example, if the first two authors
of an article contributed equally to the work.

If your author list is lengthy, you must define a shortened version of
the list of authors to be used in the page headers, to prevent
overlapping text. The following command should be placed just after
the last \verb|\author{}| definition:
\begin{verbatim}
  \renewcommand{\shortauthors}{McCartney, et al.}
\end{verbatim}
Omitting this command will force the use of a concatenated list of all
of the authors' names, which may result in overlapping text in the
page headers.

The article template's documentation, available at
\url{https://www.acm.org/publications/proceedings-template}, has a
complete explanation of these commands and tips for their effective
use.

Note that authors' addresses are mandatory for journal articles.

\section{Rights Information}

Authors of any work published by ACM will need to complete a rights
form. Depending on the kind of work, and the rights management choice
made by the author, this may be copyright transfer, permission,
license, or an OA (open access) agreement.

Regardless of the rights management choice, the author will receive a
copy of the completed rights form once it has been submitted. This
form contains \LaTeX\ commands that must be copied into the source
document. When the document source is compiled, these commands and
their parameters add formatted text to several areas of the final
document:
\begin{itemize}
\item the ``ACM Reference Format'' text on the first page.
\item the ``rights management'' text on the first page.
\item the conference information in the page header(s).
\end{itemize}

Rights information is unique to the work; if you are preparing several
works for an event, make sure to use the correct set of commands with
each of the works.

The ACM Reference Format text is required for all articles over one
page in length, and is optional for one-page articles (abstracts).

\section{CCS Concepts and User-Defined Keywords}

Two elements of the ``acmart'' document class provide powerful
taxonomic tools for you to help readers find your work in an online
search.

The ACM Computing Classification System ---
\url{https://www.acm.org/publications/class-2012} --- is a set of
classifiers and concepts that describe the computing
discipline. Authors can select entries from this classification
system, via \url{https://dl.acm.org/ccs/ccs.cfm}, and generate the
commands to be included in the \LaTeX\ source.

User-defined keywords are a comma-separated list of words and phrases
of the authors' choosing, providing a more flexible way of describing
the research being presented.

CCS concepts and user-defined keywords are required for for all
articles over two pages in length, and are optional for one- and
two-page articles (or abstracts).

\section{Sectioning Commands}

Your work should use standard \LaTeX\ sectioning commands:
\verb|section|, \verb|subsection|, \verb|subsubsection|, and
\verb|paragraph|. They should be numbered; do not remove the numbering
from the commands.

Simulating a sectioning command by setting the first word or words of
a paragraph in boldface or italicized text is {\bfseries not allowed.}

\section{Tables}

The ``\verb|acmart|'' document class includes the ``\verb|booktabs|''
package --- \url{https://ctan.org/pkg/booktabs} --- for preparing
high-quality tables.

Table captions are placed {\itshape above} the table.

Because tables cannot be split across pages, the best placement for
them is typically the top of the page nearest their initial cite.  To
ensure this proper ``floating'' placement of tables, use the
environment \textbf{table} to enclose the table's contents and the
table caption.  The contents of the table itself must go in the
\textbf{tabular} environment, to be aligned properly in rows and
columns, with the desired horizontal and vertical rules.  Again,
detailed instructions on \textbf{tabular} material are found in the
\textit{\LaTeX\ User's Guide}.

Immediately following this sentence is the point at which
Table~\ref{tab:freq} is included in the input file; compare the
placement of the table here with the table in the printed output of
this document.

\begin{table}
  \caption{Frequency of Special Characters}
  \label{tab:freq}
  \begin{tabular}{ccl}
    \toprule
    Non-English or Math&Frequency&Comments\\
    \midrule
    \O & 1 in 1,000& For Swedish names\\
    $\pi$ & 1 in 5& Common in math\\
    \$ & 4 in 5 & Used in business\\
    $\Psi^2_1$ & 1 in 40,000& Unexplained usage\\
  \bottomrule
\end{tabular}
\end{table}

To set a wider table, which takes up the whole width of the page's
live area, use the environment \textbf{table*} to enclose the table's
contents and the table caption.  As with a single-column table, this
wide table will ``float'' to a location deemed more
desirable. Immediately following this sentence is the point at which
Table~\ref{tab:commands} is included in the input file; again, it is
instructive to compare the placement of the table here with the table
in the printed output of this document.

\begin{table*}
  \caption{Some Typical Commands}
  \label{tab:commands}
  \begin{tabular}{ccl}
    \toprule
    Command &A Number & Comments\\
    \midrule
    \texttt{{\char'134}author} & 100& Author \\
    \texttt{{\char'134}table}& 300 & For tables\\
    \texttt{{\char'134}table*}& 400& For wider tables\\
    \bottomrule
  \end{tabular}
\end{table*}

Always use midrule to separate table header rows from data rows, and
use it only for this purpose. This enables assistive technologies to
recognise table headers and support their users in navigating tables
more easily.

\section{Math Equations}
You may want to display math equations in three distinct styles:
inline, numbered or non-numbered display.  Each of the three are
discussed in the next sections.

\subsection{Inline (In-text) Equations}
A formula that appears in the running text is called an inline or
in-text formula.  It is produced by the \textbf{math} environment,
which can be invoked with the usual
\texttt{{\char'134}begin\,\ldots{\char'134}end} construction or with
the short form \texttt{\$\,\ldots\$}. You can use any of the symbols
and structures, from $\alpha$ to $\omega$, available in
\LaTeX~\cite{Lamport:LaTeX}; this section will simply show a few
examples of in-text equations in context. Notice how this equation:
\begin{math}
  \lim_{n\rightarrow \infty}x=0
\end{math},
set here in in-line math style, looks slightly different when
set in display style.  (See next section).

\subsection{Display Equations}
A numbered display equation---one set off by vertical space from the
text and centered horizontally---is produced by the \textbf{equation}
environment. An unnumbered display equation is produced by the
\textbf{displaymath} environment.

Again, in either environment, you can use any of the symbols and
structures available in \LaTeX\@; this section will just give a couple
of examples of display equations in context.  First, consider the
equation, shown as an inline equation above:
\begin{equation}
  \lim_{n\rightarrow \infty}x=0
\end{equation}
Notice how it is formatted somewhat differently in
the \textbf{displaymath}
environment.  Now, we'll enter an unnumbered equation:
\begin{displaymath}
  \sum_{i=0}^{\infty} x + 1
\end{displaymath}
and follow it with another numbered equation:
\begin{equation}
  \sum_{i=0}^{\infty}x_i=\int_{0}^{\pi+2} f
\end{equation}
just to demonstrate \LaTeX's able handling of numbering.

\section{Figures}

The ``\verb|figure|'' environment should be used for figures. One or
more images can be placed within a figure. If your figure contains
third-party material, you must clearly identify it as such, as shown
in the example below.
\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{sample-franklin}
  \caption{1907 Franklin Model D roadster. Photograph by Harris \&
    Ewing, Inc. [Public domain], via Wikimedia
    Commons. (\url{https://goo.gl/VLCRBB}).}
  \Description{A woman and a girl in white dresses sit in an open car.}
\end{figure}

Your figures should contain a caption which describes the figure to
the reader.

Figure captions are placed {\itshape below} the figure.

Every figure should also have a figure description unless it is purely
decorative. These descriptions convey what’s in the image to someone
who cannot see it. They are also used by search engine crawlers for
indexing images, and when images cannot be loaded.

A figure description must be unformatted plain text less than 2000
characters long (including spaces).  {\bfseries Figure descriptions
  should not repeat the figure caption – their purpose is to capture
  important information that is not already provided in the caption or
  the main text of the paper.} For figures that convey important and
complex new information, a short text description may not be
adequate. More complex alternative descriptions can be placed in an
appendix and referenced in a short figure description. For example,
provide a data table capturing the information in a bar chart, or a
structured list representing a graph.  For additional information
regarding how best to write figure descriptions and why doing this is
so important, please see
\url{https://www.acm.org/publications/taps/describing-figures/}.

\subsection{The ``Teaser Figure''}

A ``teaser figure'' is an image, or set of images in one figure, that
are placed after all author and affiliation information, and before
the body of the article, spanning the page. If you wish to have such a
figure in your article, place the command immediately before the
\verb|\maketitle| command:
\begin{verbatim}
  \begin{teaserfigure}
    \includegraphics[width=\textwidth]{sampleteaser}
    \caption{figure caption}
    \Description{figure description}
  \end{teaserfigure}
\end{verbatim}

\section{Citations and Bibliographies}

The use of \BibTeX\ for the preparation and formatting of one's
references is strongly recommended. Authors' names should be complete
--- use full first names (``Donald E. Knuth'') not initials
(``D. E. Knuth'') --- and the salient identifying features of a
reference should be included: title, year, volume, number, pages,
article DOI, etc.

The bibliography is included in your source document with these two
commands, placed just before the \verb|\end{document}| command:
\begin{verbatim}
  \bibliographystyle{ACM-Reference-Format}
  \bibliography{bibfile}
\end{verbatim}
where ``\verb|bibfile|'' is the name, without the ``\verb|.bib|''
suffix, of the \BibTeX\ file.

Citations and references are numbered by default. A small number of
ACM publications have citations and references formatted in the
``author year'' style; for these exceptions, please include this
command in the {\bfseries preamble} (before the command
``\verb|\begin{document}|'') of your \LaTeX\ source:
\begin{verbatim}
  \citestyle{acmauthoryear}
\end{verbatim}

  Some examples.  A paginated journal article \cite{Abril07}, an
  enumerated journal article \cite{Cohen07}, a reference to an entire
  issue \cite{JCohen96}, a monograph (whole book) \cite{Kosiur01}, a
  monograph/whole book in a series (see 2a in spec. document)
  \cite{Harel79}, a divisible-book such as an anthology or compilation
  \cite{Editor00} followed by the same example, however we only output
  the series if the volume number is given \cite{Editor00a} (so
  Editor00a's series should NOT be present since it has no vol. no.),
  a chapter in a divisible book \cite{Spector90}, a chapter in a
  divisible book in a series \cite{Douglass98}, a multi-volume work as
  book \cite{Knuth97}, a couple of articles in a proceedings (of a
  conference, symposium, workshop for example) (paginated proceedings
  article) \cite{Andler79, Hagerup1993}, a proceedings article with
  all possible elements \cite{Smith10}, an example of an enumerated
  proceedings article \cite{VanGundy07}, an informally published work
  \cite{Harel78}, a couple of preprints \cite{Bornmann2019,
    AnzarootPBM14}, a doctoral dissertation \cite{Clarkson85}, a
  master's thesis: \cite{anisi03}, an online document / world wide web
  resource \cite{Thornburg01, Ablamowicz07, Poker06}, a video game
  (Case 1) \cite{Obama08} and (Case 2) \cite{Novak03} and \cite{Lee05}
  and (Case 3) a patent \cite{JoeScientist001}, work accepted for
  publication \cite{rous08}, 'YYYYb'-test for prolific author
  \cite{SaeediMEJ10} and \cite{SaeediJETC10}. Other cites might
  contain 'duplicate' DOI and URLs (some SIAM articles)
  \cite{Kirschmer:2010:AEI:1958016.1958018}. Boris / Barbara Beeton:
  multi-volume works as books \cite{MR781536} and \cite{MR781537}. A
  couple of citations with DOIs:
  \cite{2004:ITE:1009386.1010128,Kirschmer:2010:AEI:1958016.1958018}. Online
  citations: \cite{TUGInstmem, Thornburg01, CTANacmart}.
  Artifacts: \cite{R} and \cite{UMassCitations}.

\section{Acknowledgments}

Identification of funding sources and other support, and thanks to
individuals and groups that assisted in the research and the
preparation of the work should be included in an acknowledgment
section, which is placed just before the reference section in your
document.

This section has a special environment:
\begin{verbatim}
  \begin{acks}
  ...
  \end{acks}
\end{verbatim}
so that the information contained therein can be more easily collected
during the article metadata extraction phase, and to ensure
consistency in the spelling of the section heading.

Authors should not prepare this section as a numbered or unnumbered {\verb|\section|}; please use the ``{\verb|acks|}'' environment.

\section{Appendices}

If your work needs an appendix, add it before the
``\verb|\end{document}|'' command at the conclusion of your source
document.

Start the appendix with the ``\verb|appendix|'' command:
\begin{verbatim}
  \appendix
\end{verbatim}
and note that in the appendix, sections are lettered, not
numbered. This document has two appendices, demonstrating the section
and subsection identification method.

\section{Multi-language papers}

Papers may be written in languages other than English or include
titles, subtitles, keywords and abstracts in different languages (as a
rule, a paper in a language other than English should include an
English title and an English abstract).  Use \verb|language=...| for
every language used in the paper.  The last language indicated is the
main language of the paper.  For example, a French paper with
additional titles and abstracts in English and German may start with
the following command
\begin{verbatim}
\documentclass[sigconf, language=english, language=german,
               language=french]{acmart}
\end{verbatim}

The title, subtitle, keywords and abstract will be typeset in the main
language of the paper.  The commands \verb|\translatedXXX|, \verb|XXX|
begin title, subtitle and keywords, can be used to set these elements
in the other languages.  The environment \verb|translatedabstract| is
used to set the translation of the abstract.  These commands and
environment have a mandatory first argument: the language of the
second argument.  See \verb|sample-sigconf-i13n.tex| file for examples
of their usage.

\section{SIGCHI Extended Abstracts}

The ``\verb|sigchi-a|'' template style (available only in \LaTeX\ and
not in Word) produces a landscape-orientation formatted article, with
a wide left margin. Three environments are available for use with the
``\verb|sigchi-a|'' template style, and produce formatted output in
the margin:
\begin{description}
\item[\texttt{sidebar}:]  Place formatted text in the margin.
\item[\texttt{marginfigure}:] Place a figure in the margin.
\item[\texttt{margintable}:] Place a table in the margin.
\end{description}

%%
%% The acknowledgments section is defined using the "acks" environment
%% (and NOT an unnumbered section). This ensures the proper
%% identification of the section in the article metadata, and the
%% consistent spelling of the heading.
\begin{acks}
To Robert, for the bagels and explaining CMYK and color spaces.
\end{acks}

%%
%% The next two lines define the bibliography style to be used, and
%% the bibliography file.
\bibliographystyle{ACM-Reference-Format}
\bibliography{sample-base}

%%
%% If your work has an appendix, this is the place to put it.
\appendix

\section{Research Methods}

\subsection{Part One}

Lorem ipsum dolor sit amet, consectetur adipiscing elit. Morbi
malesuada, quam in pulvinar varius, metus nunc fermentum urna, id
sollicitudin purus odio sit amet enim. Aliquam ullamcorper eu ipsum
vel mollis. Curabitur quis dictum nisl. Phasellus vel semper risus, et
lacinia dolor. Integer ultricies commodo sem nec semper.

\subsection{Part Two}

Etiam commodo feugiat nisl pulvinar pellentesque. Etiam auctor sodales
ligula, non varius nibh pulvinar semper. Suspendisse nec lectus non
ipsum convallis congue hendrerit vitae sapien. Donec at laoreet
eros. Vivamus non purus placerat, scelerisque diam eu, cursus
ante. Etiam aliquam tortor auctor efficitur mattis.

\section{Online Resources}

Nam id fermentum dui. Suspendisse sagittis tortor a nulla mollis, in
pulvinar ex pretium. Sed interdum orci quis metus euismod, et sagittis
enim maximus. Vestibulum gravida massa ut felis suscipit
congue. Quisque mattis elit a risus ultrices commodo venenatis eget
dui. Etiam sagittis eleifend elementum.

Nam interdum magna at lectus dignissim, ac dignissim lorem
rhoncus. Maecenas eu arcu ac neque placerat aliquam. Nunc pulvinar
massa et mattis lacinia.

\end{document}
\endinput
%%
%% End of file `sample-acmsmall-conf.tex'.

### 2022-10-29 paper.tex
making bib
trying to get the acmart reference to work...

@@
 \input{pldi23-preamble}
+
+\usepackage{utf8math,ttquot,mathpartir,amsmath, amssymb, hydrocomments, mathtools,multicol,xspace,stmaryrd }
 
@@
 \section{Introduction}
-ACM's consolidated article template, introduced in 2017, provides a
-consistent \LaTeX\ style for use across ACM publications, and
-incorporates accessibility and metadata-extraction functionality
-necessary for future Digital Library endeavors. Numerous ACM and
-SIG-specific \LaTeX\ templates have been examined, and their unique
-features incorporated into this single new template.
+ACM's consolidated article template \cite{mixt},
 
-If you are new to publishing with ACM, this document is a valuable
-guide to the process of preparing your work for publication. If you
-have published with ACM before, this document provides insight and
-instruction into more recent changes to the article template.
-
-The ``\verb|acmart|'' document class can be used to prepare articles
-for any ACM publication --- conference or journal, and for any stage
-of publication, from review to final ``camera-ready'' copy, to the
-author's own version, with {\itshape very} few changes to the source.
-
-\section{Template Overview}
-As noted in the introduction, the ``\verb|acmart|'' document class can
-be used to prepare many different kinds of documentation --- a
-double-blind initial submission of a full-length technical paper, a
-two-page SIGGRAPH Emerging Technologies abstract, a ``camera-ready''
-journal article, a SIGCHI Extended Abstract, and more --- all by
-selecting the appropriate {\itshape template style} and {\itshape
-  template parameters}.
-
-This document will explain the major features of the document
-class. For further information, the {\itshape \LaTeX\ User's Guide} is
-available from
-\url{https://www.acm.org/publications/proceedings-template}.
-
-\subsection{Template Styles}
-
-The primary parameter given to the ``\verb|acmart|'' document class is
-the {\itshape template style} which corresponds to the kind of publication
-or SIG publishing the work. This parameter is enclosed in square
-brackets and is a part of the {\verb|documentclass|} command:
-\begin{verbatim}
-  \documentclass[STYLE]{acmart}
-\end{verbatim}
-
-Journals use one of three template styles. All but three ACM journals
-use the {\verb|acmsmall|} template style:
-\begin{itemize}
-\item {\texttt{acmsmall}}: The default journal template style.
-\item {\texttt{acmlarge}}: Used by JOCCH and TAP.
-\item {\texttt{acmtog}}: Used by TOG.
-\end{itemize}
-
-The majority of conference proceedings documentation will use the {\verb|acmconf|} template style.
-\begin{itemize}
-\item {\texttt{acmconf}}: The default proceedings template style.
-\item{\texttt{sigchi}}: Used for SIGCHI conference articles.
-\item{\texttt{sigplan}}: Used for SIGPLAN conference articles.
-\end{itemize}
-
-\subsection{Template Parameters}
-
-In addition to specifying the {\itshape template style} to be used in
-formatting your work, there are a number of {\itshape template parameters}
-which modify some part of the applied template style. A complete list
-of these parameters can be found in the {\itshape \LaTeX\ User's Guide.}
-
-Frequently-used parameters, or combinations of parameters, include:
-\begin{itemize}
-\item {\texttt{anonymous,review}}: Suitable for a ``double-blind''
-  conference submission. Anonymizes the work and includes line
-  numbers. Use with the \texttt{\acmSubmissionID} command to print the
-  submission's unique ID on each page of the work.
-\item{\texttt{authorversion}}: Produces a version of the work suitable
-  for posting by the author.
-\item{\texttt{screen}}: Produces colored hyperlinks.
-\end{itemize}
-
-This document uses the following string as the first command in the
-source file:
-\begin{verbatim}
-\documentclass[acmsmall]{acmart}
-\end{verbatim}
-
-\section{Modifications}
-
-Modifying the template --- including but not limited to: adjusting
-margins, typeface sizes, line spacing, paragraph and list definitions,
-and the use of the \verb|\vspace| command to manually adjust the
-vertical spacing between elements of your work --- is not allowed.
-
-{\bfseries Your document will be returned to you for revision if
-  modifications are discovered.}
-
-\section{Typefaces}
-
-The ``\verb|acmart|'' document class requires the use of the
-``Libertine'' typeface family. Your \TeX\ installation should include
-this set of packages. Please do not substitute other typefaces. The
-``\verb|lmodern|'' and ``\verb|ltimes|'' packages should not be used,
-as they will override the built-in typeface families.
-
-\section{Title Information}
-
-The title of your work should use capital letters appropriately -
-\url{https://capitalizemytitle.com/} has useful rules for
-capitalization. Use the {\verb|title|} command to define the title of
-your work. If your work has a subtitle, define it with the
-{\verb|subtitle|} command.  Do not insert line breaks in your title.
-
-If your title is lengthy, you must define a short version to be used
-in the page headers, to prevent overlapping text. The \verb|title|
-command has a ``short title'' parameter:
-\begin{verbatim}
-  \title[short title]{full title}
-\end{verbatim}
-
-\section{Authors and Affiliations}
-
-Each author must be defined separately for accurate metadata
-identification.  As an exception, multiple authors may share one
-affiliation. Authors' names should not be abbreviated; use full first
-names wherever possible. Include authors' e-mail addresses whenever
-possible.
-
-Grouping authors' names or e-mail addresses, or providing an ``e-mail
-alias,'' as shown below, is not acceptable:
-\begin{verbatim}
-  \author{Brooke Aster, David Mehldau}
-  \email{dave,judy,steve@university.edu}
-  \email{firstname.lastname@phillips.org}
-\end{verbatim}
-
-The \verb|authornote| and \verb|authornotemark| commands allow a note
-to apply to multiple authors --- for example, if the first two authors
-of an article contributed equally to the work.
-
-If your author list is lengthy, you must define a shortened version of
-the list of authors to be used in the page headers, to prevent
-overlapping text. The following command should be placed just after
-the last \verb|\author{}| definition:
-\begin{verbatim}
-  \renewcommand{\shortauthors}{McCartney, et al.}
-\end{verbatim}
-Omitting this command will force the use of a concatenated list of all
-of the authors' names, which may result in overlapping text in the
-page headers.
-
-The article template's documentation, available at
-\url{https://www.acm.org/publications/proceedings-template}, has a
-complete explanation of these commands and tips for their effective
-use.
-
-Note that authors' addresses are mandatory for journal articles.
-
-\section{Rights Information}
-
-Authors of any work published by ACM will need to complete a rights
-form. Depending on the kind of work, and the rights management choice
-made by the author, this may be copyright transfer, permission,
-license, or an OA (open access) agreement.
-
-Regardless of the rights management choice, the author will receive a
-copy of the completed rights form once it has been submitted. This
-form contains \LaTeX\ commands that must be copied into the source
-document. When the document source is compiled, these commands and
-their parameters add formatted text to several areas of the final
-document:
-\begin{itemize}
-\item the ``ACM Reference Format'' text on the first page.
-\item the ``rights management'' text on the first page.
-\item the conference information in the page header(s).
-\end{itemize}
-
-Rights information is unique to the work; if you are preparing several
-works for an event, make sure to use the correct set of commands with
-each of the works.
-
-The ACM Reference Format text is required for all articles over one
-page in length, and is optional for one-page articles (abstracts).
-
-\section{CCS Concepts and User-Defined Keywords}
-
-Two elements of the ``acmart'' document class provide powerful
-taxonomic tools for you to help readers find your work in an online
-search.
-
-The ACM Computing Classification System ---
-\url{https://www.acm.org/publications/class-2012} --- is a set of
-classifiers and concepts that describe the computing
-discipline. Authors can select entries from this classification
-system, via \url{https://dl.acm.org/ccs/ccs.cfm}, and generate the
-commands to be included in the \LaTeX\ source.
-
-User-defined keywords are a comma-separated list of words and phrases
-of the authors' choosing, providing a more flexible way of describing
-the research being presented.
-
-CCS concepts and user-defined keywords are required for for all
-articles over two pages in length, and are optional for one- and
-two-page articles (or abstracts).
-
-\section{Sectioning Commands}
-
-Your work should use standard \LaTeX\ sectioning commands:
-\verb|section|, \verb|subsection|, \verb|subsubsection|, and
-\verb|paragraph|. They should be numbered; do not remove the numbering
-from the commands.
-
-Simulating a sectioning command by setting the first word or words of
-a paragraph in boldface or italicized text is {\bfseries not allowed.}
-
-\section{Tables}
-
-The ``\verb|acmart|'' document class includes the ``\verb|booktabs|''
-package --- \url{https://ctan.org/pkg/booktabs} --- for preparing
-high-quality tables.
-
-Table captions are placed {\itshape above} the table.
-
-Because tables cannot be split across pages, the best placement for
-them is typically the top of the page nearest their initial cite.  To
-ensure this proper ``floating'' placement of tables, use the
-environment \textbf{table} to enclose the table's contents and the
-table caption.  The contents of the table itself must go in the
-\textbf{tabular} environment, to be aligned properly in rows and
-columns, with the desired horizontal and vertical rules.  Again,
-detailed instructions on \textbf{tabular} material are found in the
-\textit{\LaTeX\ User's Guide}.
-
-Immediately following this sentence is the point at which
-Table~\ref{tab:freq} is included in the input file; compare the
-placement of the table here with the table in the printed output of
-this document.
-
-\begin{table}
-  \caption{Frequency of Special Characters}
-  \label{tab:freq}
-  \begin{tabular}{ccl}
-    \toprule
-    Non-English or Math&Frequency&Comments\\
-    \midrule
-    \O & 1 in 1,000& For Swedish names\\
-    $\pi$ & 1 in 5& Common in math\\
-    \$ & 4 in 5 & Used in business\\
-    $\Psi^2_1$ & 1 in 40,000& Unexplained usage\\
-  \bottomrule
-\end{tabular}
-\end{table}
-
-To set a wider table, which takes up the whole width of the page's
-live area, use the environment \textbf{table*} to enclose the table's
-contents and the table caption.  As with a single-column table, this
-wide table will ``float'' to a location deemed more
-desirable. Immediately following this sentence is the point at which
-Table~\ref{tab:commands} is included in the input file; again, it is
-instructive to compare the placement of the table here with the table
-in the printed output of this document.
-
-\begin{table*}
-  \caption{Some Typical Commands}
-  \label{tab:commands}
-  \begin{tabular}{ccl}
-    \toprule
-    Command &A Number & Comments\\
-    \midrule
-    \texttt{{\char'134}author} & 100& Author \\
-    \texttt{{\char'134}table}& 300 & For tables\\
-    \texttt{{\char'134}table*}& 400& For wider tables\\
-    \bottomrule
-  \end{tabular}
-\end{table*}
-
-Always use midrule to separate table header rows from data rows, and
-use it only for this purpose. This enables assistive technologies to
-recognise table headers and support their users in navigating tables
-more easily.
-
-\section{Math Equations}
-You may want to display math equations in three distinct styles:
-inline, numbered or non-numbered display.  Each of the three are
-discussed in the next sections.
-
-\subsection{Inline (In-text) Equations}
-A formula that appears in the running text is called an inline or
-in-text formula.  It is produced by the \textbf{math} environment,
-which can be invoked with the usual
-\texttt{{\char'134}begin\,\ldots{\char'134}end} construction or with
-the short form \texttt{\$\,\ldots\$}. You can use any of the symbols
-and structures, from $\alpha$ to $\omega$, available in
-\LaTeX~\cite{Lamport:LaTeX}; this section will simply show a few
-examples of in-text equations in context. Notice how this equation:
-\begin{math}
-  \lim_{n\rightarrow \infty}x=0
-\end{math},
-set here in in-line math style, looks slightly different when
-set in display style.  (See next section).
-
-\subsection{Display Equations}
-A numbered display equation---one set off by vertical space from the
-text and centered horizontally---is produced by the \textbf{equation}
-environment. An unnumbered display equation is produced by the
-\textbf{displaymath} environment.
-
-Again, in either environment, you can use any of the symbols and
-structures available in \LaTeX\@; this section will just give a couple
-of examples of display equations in context.  First, consider the
-equation, shown as an inline equation above:
-\begin{equation}
-  \lim_{n\rightarrow \infty}x=0
-\end{equation}
-Notice how it is formatted somewhat differently in
-the \textbf{displaymath}
-environment.  Now, we'll enter an unnumbered equation:
-\begin{displaymath}
-  \sum_{i=0}^{\infty} x + 1
-\end{displaymath}
-and follow it with another numbered equation:
-\begin{equation}
-  \sum_{i=0}^{\infty}x_i=\int_{0}^{\pi+2} f
-\end{equation}
-just to demonstrate \LaTeX's able handling of numbering.
-
-\section{Figures}
-
-The ``\verb|figure|'' environment should be used for figures. One or
-more images can be placed within a figure. If your figure contains
-third-party material, you must clearly identify it as such, as shown
-in the example below.
-\begin{figure}[h]
-  \centering
-  \includegraphics[width=\linewidth]{sample-franklin}
-  \caption{1907 Franklin Model D roadster. Photograph by Harris \&
-    Ewing, Inc. [Public domain], via Wikimedia
-    Commons. (\url{https://goo.gl/VLCRBB}).}
-  \Description{A woman and a girl in white dresses sit in an open car.}
-\end{figure}
-
-Your figures should contain a caption which describes the figure to
-the reader.
-
-Figure captions are placed {\itshape below} the figure.
-
-Every figure should also have a figure description unless it is purely
-decorative. These descriptions convey what’s in the image to someone
-who cannot see it. They are also used by search engine crawlers for
-indexing images, and when images cannot be loaded.
-
-A figure description must be unformatted plain text less than 2000
-characters long (including spaces).  {\bfseries Figure descriptions
-  should not repeat the figure caption – their purpose is to capture
-  important information that is not already provided in the caption or
-  the main text of the paper.} For figures that convey important and
-complex new information, a short text description may not be
-adequate. More complex alternative descriptions can be placed in an
-appendix and referenced in a short figure description. For example,
-provide a data table capturing the information in a bar chart, or a
-structured list representing a graph.  For additional information
-regarding how best to write figure descriptions and why doing this is
-so important, please see
-\url{https://www.acm.org/publications/taps/describing-figures/}.
-
-\subsection{The ``Teaser Figure''}
-
-A ``teaser figure'' is an image, or set of images in one figure, that
-are placed after all author and affiliation information, and before
-the body of the article, spanning the page. If you wish to have such a
-figure in your article, place the command immediately before the
-\verb|\maketitle| command:
-\begin{verbatim}
-  \begin{teaserfigure}
-    \includegraphics[width=\textwidth]{sampleteaser}
-    \caption{figure caption}
-    \Description{figure description}
-  \end{teaserfigure}
-\end{verbatim}
-
-\section{Citations and Bibliographies}
-
-The use of \BibTeX\ for the preparation and formatting of one's
-references is strongly recommended. Authors' names should be complete
---- use full first names (``Donald E. Knuth'') not initials
-(``D. E. Knuth'') --- and the salient identifying features of a
-reference should be included: title, year, volume, number, pages,
-article DOI, etc.
-
-The bibliography is included in your source document with these two
-commands, placed just before the \verb|\end{document}| command:
-\begin{verbatim}
 \bibliographystyle{ACM-Reference-Format}
-  \bibliography{bibfile}
-\end{verbatim}
-where ``\verb|bibfile|'' is the name, without the ``\verb|.bib|''
-suffix, of the \BibTeX\ file.
-
-Citations and references are numbered by default. A small number of
-ACM publications have citations and references formatted in the
-``author year'' style; for these exceptions, please include this
-command in the {\bfseries preamble} (before the command
-``\verb|\begin{document}|'') of your \LaTeX\ source:
-\begin{verbatim}
 \citestyle{acmauthoryear}
-\end{verbatim}
-
-  Some examples.  A paginated journal article \cite{Abril07}, an
-  enumerated journal article \cite{Cohen07}, a reference to an entire
-  issue \cite{JCohen96}, a monograph (whole book) \cite{Kosiur01}, a
-  monograph/whole book in a series (see 2a in spec. document)
-  \cite{Harel79}, a divisible-book such as an anthology or compilation
-  \cite{Editor00} followed by the same example, however we only output
-  the series if the volume number is given \cite{Editor00a} (so
-  Editor00a's series should NOT be present since it has no vol. no.),
-  a chapter in a divisible book \cite{Spector90}, a chapter in a
-  divisible book in a series \cite{Douglass98}, a multi-volume work as
-  book \cite{Knuth97}, a couple of articles in a proceedings (of a
-  conference, symposium, workshop for example) (paginated proceedings
-  article) \cite{Andler79, Hagerup1993}, a proceedings article with
-  all possible elements \cite{Smith10}, an example of an enumerated
-  proceedings article \cite{VanGundy07}, an informally published work
-  \cite{Harel78}, a couple of preprints \cite{Bornmann2019,
-    AnzarootPBM14}, a doctoral dissertation \cite{Clarkson85}, a
-  master's thesis: \cite{anisi03}, an online document / world wide web
-  resource \cite{Thornburg01, Ablamowicz07, Poker06}, a video game
-  (Case 1) \cite{Obama08} and (Case 2) \cite{Novak03} and \cite{Lee05}
-  and (Case 3) a patent \cite{JoeScientist001}, work accepted for
-  publication \cite{rous08}, 'YYYYb'-test for prolific author
-  \cite{SaeediMEJ10} and \cite{SaeediJETC10}. Other cites might
-  contain 'duplicate' DOI and URLs (some SIAM articles)
-  \cite{Kirschmer:2010:AEI:1958016.1958018}. Boris / Barbara Beeton:
-  multi-volume works as books \cite{MR781536} and \cite{MR781537}. A
-  couple of citations with DOIs:
-  \cite{2004:ITE:1009386.1010128,Kirschmer:2010:AEI:1958016.1958018}. Online
-  citations: \cite{TUGInstmem, Thornburg01, CTANacmart}.
-  Artifacts: \cite{R} and \cite{UMassCitations}.
-
-\section{Acknowledgments}
-
-Identification of funding sources and other support, and thanks to
-individuals and groups that assisted in the research and the
-preparation of the work should be included in an acknowledgment
-section, which is placed just before the reference section in your
-document.
-
-This section has a special environment:
-\begin{verbatim}
-  \begin{acks}
-  ...
-  \end{acks}
-\end{verbatim}
-so that the information contained therein can be more easily collected
-during the article metadata extraction phase, and to ensure
-consistency in the spelling of the section heading.
-
-Authors should not prepare this section as a numbered or unnumbered {\verb|\section|}; please use the ``{\verb|acks|}'' environment.
-
-\section{Appendices}
-
-If your work needs an appendix, add it before the
-``\verb|\end{document}|'' command at the conclusion of your source
-document.
-
-Start the appendix with the ``\verb|appendix|'' command:
-\begin{verbatim}
-  \appendix
-\end{verbatim}
-and note that in the appendix, sections are lettered, not
-numbered. This document has two appendices, demonstrating the section
-and subsection identification method.
-
-\section{Multi-language papers}
-
-Papers may be written in languages other than English or include
-titles, subtitles, keywords and abstracts in different languages (as a
-rule, a paper in a language other than English should include an
-English title and an English abstract).  Use \verb|language=...| for
-every language used in the paper.  The last language indicated is the
-main language of the paper.  For example, a French paper with
-additional titles and abstracts in English and German may start with
-the following command
-\begin{verbatim}
-\documentclass[sigconf, language=english, language=german,
-               language=french]{acmart}
-\end{verbatim}
-
-The title, subtitle, keywords and abstract will be typeset in the main
-language of the paper.  The commands \verb|\translatedXXX|, \verb|XXX|
-begin title, subtitle and keywords, can be used to set these elements
-in the other languages.  The environment \verb|translatedabstract| is
-used to set the translation of the abstract.  These commands and
-environment have a mandatory first argument: the language of the
-second argument.  See \verb|sample-sigconf-i13n.tex| file for examples
-of their usage.
-
-\section{SIGCHI Extended Abstracts}
-
-The ``\verb|sigchi-a|'' template style (available only in \LaTeX\ and
-not in Word) produces a landscape-orientation formatted article, with
-a wide left margin. Three environments are available for use with the
-``\verb|sigchi-a|'' template style, and produce formatted output in
-the margin:
-\begin{description}
-\item[\texttt{sidebar}:]  Place formatted text in the margin.
-\item[\texttt{marginfigure}:] Place a figure in the margin.
-\item[\texttt{margintable}:] Place a table in the margin.
-\end{description}
-
-%%
-%% The acknowledgments section is defined using the "acks" environment
-%% (and NOT an unnumbered section). This ensures the proper
-%% identification of the section in the article metadata, and the
-%% consistent spelling of the heading.
-\begin{acks}
-To Robert, for the bagels and explaining CMYK and color spaces.
-\end{acks}
-
-%%
-%% The next two lines define the bibliography style to be used, and
-%% the bibliography file.
-\bibliographystyle{ACM-Reference-Format}
-\bibliography{sample-base}
-
-%%
-%% If your work has an appendix, this is the place to put it.
-\appendix
-
-\section{Research Methods}
-
-\subsection{Part One}
-
-Lorem ipsum dolor sit amet, consectetur adipiscing elit. Morbi
-malesuada, quam in pulvinar varius, metus nunc fermentum urna, id
-sollicitudin purus odio sit amet enim. Aliquam ullamcorper eu ipsum
-vel mollis. Curabitur quis dictum nisl. Phasellus vel semper risus, et
-lacinia dolor. Integer ultricies commodo sem nec semper.
-
-\subsection{Part Two}
-
-Etiam commodo feugiat nisl pulvinar pellentesque. Etiam auctor sodales
-ligula, non varius nibh pulvinar semper. Suspendisse nec lectus non
-ipsum convallis congue hendrerit vitae sapien. Donec at laoreet
-eros. Vivamus non purus placerat, scelerisque diam eu, cursus
-ante. Etiam aliquam tortor auctor efficitur mattis.
-
-\section{Online Resources}
-
-Nam id fermentum dui. Suspendisse sagittis tortor a nulla mollis, in
-pulvinar ex pretium. Sed interdum orci quis metus euismod, et sagittis
-enim maximus. Vestibulum gravida massa ut felis suscipit
-congue. Quisque mattis elit a risus ultrices commodo venenatis eget
-dui. Etiam sagittis eleifend elementum.
-
-Nam interdum magna at lectus dignissim, ac dignissim lorem
-rhoncus. Maecenas eu arcu ac neque placerat aliquam. Nunc pulvinar
-massa et mattis lacinia.
+\bibliography{thesis,pm-master}
 

### 2022-10-30 paper.tex
i feel insane
a few small things later, and overleaf is finally happy enough.
trying to write some "Bad" text just to have *something* in this document.

@@
-\input{pldi23-preamble}
+\input{meta/pldi23-preamble}
 
-\usepackage{utf8math,ttquot,mathpartir,amsmath, amssymb, hydrocomments, mathtools,multicol,xspace,stmaryrd }
+\let\Bbbk\relax
+\usepackage{amssymb}
+\usepackage[checkNoComments]{hydrocomments}
+\usepackage{utf8math,ttquot,mathpartir,amsmath, mathtools,multicol,xspace,stmaryrd }
 
@@
 
-\input{authors}
+\input{meta/authors}
+\input{meta/ccs-and-keywords.tex}
 
 %%
-%% The abstract is a short summary of the work to be presented in the
+%% The abstract is a short summary of tThe work to be presented in the
 %% article.
 \begin{abstract}
-  A clear and well-documented \LaTeX\ document is presented as an
-  article formatted for publication by ACM in a conference proceedings
-  or journal publication. Based on the ``acmart'' document class, this
-  article presents and explains many of the common variations, as well
-  as many of the formatting elements an author may use in the
-  preparation of the documentation of their work.
+
+%In this paper we introduce the HydroFlow programming language, a new
+
 \end{abstract}
 
-%% A "teaser" image appears between the author and affiliation
-%% information and the body of the document, and typically spans the
-%% page.
-%\begin{teaserfigure}
-%  \includegraphics[width=\textwidth]{sampleteaser}
-%  \caption{Seattle Mariners at Spring Training, 2010.}
-%  \Description{Enjoying the baseball game from the third-base
-%  seats. Ichiro Suzuki preparing to bat.}
-%  \label{fig:teaser}
-%\end{teaserfigure}
-
-%\received{20 February 2007}
-%\received[revised]{12 March 2009}
-%\received[accepted]{5 June 2009}
-
-%%
-%% This command processes the author and affiliation and title
-%% information and builds the first part of the formatted document.
 \maketitle
@@
 \section{Introduction}
-ACM's consolidated article template \cite{mixt},
+
+\MPM{This is intentionally badly-written. Think of it like an outline that was formatted like a paper.}
+
+Dataflow programming remains a compelling abstraction for cloud-based distributed computation.
+In a dataflow program, a series of computational units known as _operators_ are arranged in a graph, with output from one operator ``flowing'' to its successor.
+Each operator then processes a stream of inputs, and produces a stream of outputs directed at its next operator in sequence.
+This model of computation first gained popularity in the distributed systems community via projects like Dryad \cite{dryad} and MapReduce \cite{mapreduce}, but finds its origins within the databases community, primarily as a way of representing and executing SQL queries \cite{sql-dataflow}.
+Modern incarnations of distributed dataflow programming models, such as Materialize's Differential Dataflow and Apache's Spark and Storm \cite{spark, storm}, suffer from three major limitations: (1) Run-time type failures, (2) frequent barriers or retractions, (3) poor scalability.\MPM{Obviously we need to say more about these here, and not just wait until the background section.}
+
+In this work we present the HydroFlow programming language, a new dataflow language and runtime designed to overcome these three limitations.
+
+\section{Background: The three major problems of existing streaming frameworks}
+
+It could perhaps make sense to put a more step-by-step overview of distributed streaming systems here, rather than rely on a drive-by explanation in the intro.
+Especially considering our audience is PL, who have certainly _heard_ of these things but have no real idea how the work or why they are used.
+
+\subsection{Stream Property Violations}
+
+While the use of static types to eliminate the most common classes of errors is common in some distributed dataflow frameworks, these static types reflect only the power of the underlying type system.
+While they are able to reason about the type of the individual elements flowing through each stream, they are unable to reason about properties of the _stream itself_, which may be assumed to hold by downstream operators.
+Thus, each operators must independently enforce any stream properties at run-time, choosing to expose the programmer to crashes---or worse, logic errors.
+
+\subsection{Barriers, Retractions, and State: oh my!}
+
+When streaming computations are paired with state (as is their common case), thorny issues of consistency and determinism can quickly arise.
+Consider, for example, an operator that runs a single query φ on the entire stream as a single unit, producing a single boolean result ($φ : stream → bool$).
+Implementing such operators are challenging; existing systems generally take one of three approaches.
+Industry-standard dataflow environments (like Spark and such) take the conservative approach; they wait for a distinguished end-of-stream message\MPM{say something intelligent about batching here or something}, after which they are able to safely run φ on the entire stream.
+Other systems, most notable differential dataflow, take a different approach; they will continuously re-evaluate φ on each input, issuing a _retraction_ if the result of φ should ever change.
+It is then the responsibility of downstream operators to determine how to use this retraction to adjust their own computational results.
+Within the PL space, reactive programming frameworks have solved this problem in a number of ways; most commonly, upon the arrival of an updated value, they re-evaluate all functions which depended on that value, using a memoization strategy to avoid performing truly redundant computation.
+
+All of these solutions have significant drawbacks.
+In distributed dataflow, a batching approach significantly complicates the problem of distributed scheduling; downstream operators must sit idle while whole-stream queries wait for their inputs to finish arriving, and so great effort must be spent to ensure that any nodes which power downstream operators are put to good use in the interim.
+Regardless of whether the scheduling problem can be solved satisfactorily, there is the issue of latency; blocking until the end of input threatens liveness, and can delay computation results indefinitely.
+
+Meanwhile, both retractions and re-executions feature significant wasted computation.
+In the best case, retractions must effectively produce _new_ computation to undo the effects of old, resulting in many CPU-hours wasted; in the worst case, retractions arrive too late, opening the application up to consistency issues.
+Re-executions, meanwhile, are unable to determine ahead of time if the re-execution will meaningfully alter the value ultimately computed by the reactive program, stopping only when a cached value has been dynamically detected.
+This can lead to significant wasted re-computation; for example, when computing if the cardinality of a set is above a certain threshold, these computations would continuously re-compute the set cardinality, only to determine that it has no effect on the threshold and thus no effect on the overall computation.
+
+\subsection{Scaling manually, if at all}
 

### 2022-10-31 paper.tex
a forwarding pointer

@@
 
+If you're looking for the working document with all my mathy notes about the type system, head over to "notes/types.tex"
+
 \end{abstract}
@@
 In this work we present the HydroFlow programming language, a new dataflow language and runtime designed to overcome these three limitations.
+I should say a sentence here about the basics of HydroFlow, like users write Operators in a traditional systems programming language, and combine operators into stream-processing programs via a simple graph description syntax.
+First, HydroFlow's strong static type system is capable of encoding not just the types of data elements that flow through streams, but also _properties_ of the streams themselves: for example, the type of items in a hydroflow stream might statically indicate that stream elements are comparable, while the _stream properties_ may ensure that these elements arrive in _monotonically ascending_ order.
+This strong type system can statically detect many more type errors, preventing run-time crashes.
+Second, HydroFlow can leverage its stream properties to perform a _monotonicity analysis_, indicating whether a given operator is safe to run _incrementally_, and thereby solving the latency issues encountered by the introduction of barriers in traditional stream processing systems, or the consistency issues which arise from the unwise elision thereof.
+Finally, HydroFlow addresses dataflow scalability concerns by introducing rewriting-based optimizations familiar to the world of SQL directly into the dataflow language itself.
+In HydroFlow, users may provide new operators along with an _axiomatic specification_ of their behavior; the HydroFlow compiler than uses an SMT-based analysis to determine potential rewrite axioms that relate all operators, allowing a traditional SQL-style optimization approach to full HydroFlow programs---even in the presence of arbitrary user-provided code.
+
+Hydroflow is implemented as a deep embedding into the Rust programming language \cite{rust}.
+Our compilation stack ensures all operators are compiled alongside their assembled graph, allowing the Rust compiler to perform aggressive inlining and copy-elision, erasing the overhead of  operator handoffs and greatly improving on the performance of similar tools such as timely dataflow \cite{timely-dataflow}.
+In particular, our tool demonstrates some quantifiable improvement in some benchmark or other that will prove quite compelling here.
+
+The rest of the paper proceeds as follows.  Section \ref{sec:background}
+ provides background on the challenges faced by existing distributed dataflow environments.
+Section \ref{sec:hydroflow-overview} provides an overview of the hydroflow programming language, with examples of both hydroflow programs and implementing operators.
+Section \ref{sec:hydroflow-semantics} describes the formal semantics of hydroflow, with a focus on the semantics of stream properties.
+Section \ref{sec:hydroflow-types} describes the HydroFlow type system, Section \ref{sec:hydroflow-implementation} describes the implementation, and \ref{sec:experimental-results} reviews its performance on several benchmarks.
+We also have a related work and conclusion somewhere.
 
 \section{Background: The three major problems of existing streaming frameworks}
+\label{sec:background}
 
@@
 When streaming computations are paired with state (as is their common case), thorny issues of consistency and determinism can quickly arise.
-Consider, for example, an operator that runs a single query φ on the entire stream as a single unit, producing a single boolean result ($φ : stream → bool$).
-Implementing such operators are challenging; existing systems generally take one of three approaches.
+Consider, for example, an operator that runs a single query φ on the entire stream, producing a single boolean result ($φ : stream → bool$).
+Implementing such operators is challenging; existing systems generally take one of three approaches.
 Industry-standard dataflow environments (like Spark and such) take the conservative approach; they wait for a distinguished end-of-stream message\MPM{say something intelligent about batching here or something}, after which they are able to safely run φ on the entire stream.
-Other systems, most notable differential dataflow, take a different approach; they will continuously re-evaluate φ on each input, issuing a _retraction_ if the result of φ should ever change.
+Other systems, most notably differential dataflow, take a different approach; they will continuously re-evaluate φ on each input, issuing a _retraction_ if the result of φ should ever change.
 It is then the responsibility of downstream operators to determine how to use this retraction to adjust their own computational results.
-Within the PL space, reactive programming frameworks have solved this problem in a number of ways; most commonly, upon the arrival of an updated value, they re-evaluate all functions which depended on that value, using a memoization strategy to avoid performing truly redundant computation.
+Within the PL space, reactive programming frameworks have solved this problem in a number of ways; most commonly, upon the arrival of an updated value, a reactive pipeline will be re-run up until a previously-computed intermediate result is encountered.
 
@@
 In the best case, retractions must effectively produce _new_ computation to undo the effects of old, resulting in many CPU-hours wasted; in the worst case, retractions arrive too late, opening the application up to consistency issues.
-Re-executions, meanwhile, are unable to determine ahead of time if the re-execution will meaningfully alter the value ultimately computed by the reactive program, stopping only when a cached value has been dynamically detected.
-This can lead to significant wasted re-computation; for example, when computing if the cardinality of a set is above a certain threshold, these computations would continuously re-compute the set cardinality, only to determine that it has no effect on the threshold and thus no effect on the overall computation.
+Re-executions, meanwhile, are unable to determine ahead of time if the re-execution will meaningfully alter the value computed by the reactive program, stopping only if a cached intermediate result has been dynamically detected.\MPM{I feel like I need to read more FRP to make sure this is actually true, and I shouldn't be re-pitching this as "bringing an FRP insight to dataflow" or something.}  This can lead to significant wasted re-computation; for example, when computing if the cardinality of a set is above a certain threshold, these computations would continuously re-compute the set cardinality, only to determine that it has no effect on the threshold and thus no effect on the overall computation.
 
 \subsection{Scaling manually, if at all}
+
+Distributed frameworks are often lauded for their scalability; indeed, the ability to schedule various operators across and entire clusters of nodes, rather than have fixed roles for each machine, produces a degree of dynamic flexibility and allows architectures to scale the resources provided to each individual operator.
+But these scaling approaches can only go so far; they rely on finding different execution strategies for the same graph, but ignore opportunities for optimization which require rewriting the graph itself.
+
+SQL optimizers, meanwhile, have long relied on the ability to rewrite dataflow graphs.
+A SQL query engine generates a dataflow graph from a SQL query, and can reason equationally about SQL queries in order to discover a distinct dataflow program which is capable of more efficiently computing the desired result.
+While this often produces a strictly-more-optimal query, SQL engines (unlike traditional compiler optimizations) have the benefit of knowing something about the data on which the program will be running; they can thus make optimization decisions which are tailored to the expected data, and thus can achieve far greater efficiency.
+
+Bringing these benefits to the general domain of distributed dataflow programming poses a challenge: SQL optimizers assume complete knowledge of program text, and allow only a small set of operators.
+Crucially, there are no abstractions in these programs beyond tables and no extensibility or modularity beyond ``user-defined functions (UDFs),'' which must be treated conservatively by the optimizer and which allow only a restrictive class of additional computations.
+In contrast, the world of distributed dataflow programming frequently features custom operators, and increasingly relies on abstractions in the form of both libraries and modular program design \cite{hope-this-is-true!}.
+Thus, modern dataflow libraries largely make no attempt to perform rewriting optimizations; those which do perform rewriting optimizations simply defer to SQL as an input language, allowing traditional query optimization assumptions to hold.
 

### 2022-10-31 paper.tex
Because it matters to Joe, let's throw some almost-real-looking text into a PLDI-formatted document!
wip

@@
 
+\section{An overview of the HydroFlow language}
+\label{sec:hydroflow-overview}
+
+Our basic goals for this section are to say:
+
+\begin{itemize}
+\item Hydroflow is a streaming programming language that transfers lattice elements
+\item HydroFlow assembles operators (as Rust functions) into a dataflow graph using a surface syntax
+\item HydroFlow makes independently-scheduled units, which we call _compiled components_, explicit; the surface syntax figures this out for you though.
+\item Some amount of properties?  
+\end{itemize}
+
+We now review the syntax and basic user-facing features of HydroFlow, including its programming model, extensible operators, and the user-facing portions of its type system.
+
+HydroFlow is a stream-oriented programming language, deeply embedded within Rust.
+A HydroFlow program consists of a sequence of _operator_ definitions, followed by a description of a _dataflow graph_ which assemble instances of these operators into a dataflow computation.
+
 \bibliographystyle{ACM-Reference-Format}

### 2022-10-31 paper.tex
i continue to dislike this text

--- 2022-10-31 paper.tex
@@
 
+\section{An overview of the HydroFlow language}
+\label{sec:hydroflow-overview}
+
+\subsection{Hydroflow as a dataflow library}
+
+At its most basic, hydroflow functions as a traditional dataflow
+library; it applies a sequence of operators to some stream,
+transforming a stream of inputs into a stream of outputs. As with
+modern stream-oriented libraries, Hydroflow's notion of streams is
+highly flexible; operators, which may be built-in or user-supplied,
+are able to consume and produce any number of streams (as inputs and
+outputs respectively), and may maintain internal state.
+
+\subsubsection{Defining Hydroflow operators}
+
+Operationally, HydroFlow operators are defined as ordinary Rust
+functions which take any number of arguments, each of which
+corresponds to a single element from the operator's input stream(s).
+
+...I actually don't remember the details of this part, time to read the hydroflow book.
+
+Our basic goals for this section are to say:
+
+\subsubsection{Assembling a Hydroflow Graph}
+
+With operators in hand, users may now assemble a hydroflow graph.  To
+do this, we have supplied a convenient high-level graph assembly
+syntax reminiscent of XXX, whose formal definition can be found in
+figure \ref{fig:hydroflow-pipeline-syntax}. With this syntax, users
+explicitly create instances of previously-defined operators, and
+``wire'' them together via the "->" notation.  For operators with
+multiple input or output streams, users can index operators via the
+"[n]op" (for inputs) or "op[n]" (for outputs) syntax.  Our
+graph-assembly compiler ensures that all inputs and outputs to
+operators are successfully connected, and that no output (or input) is
+used twice.
+
+\subsubsection{Running a hydroflow program}
+
+I'm not sure there's a ton to say about this, but here's a stub.
+
+\subsection{Hydroflow as a stream-oriented programming language}
+
+When viewed as a language, Hydroflow's power and promise above
+traditional dataflow libraries reveals itself. Most interesting are
+two key points: hydroflow programs as transformers for streams of
+lattice points, and Hydroflow's ability to reason about _properties_
+of these streams.
+
+\subsubsection{Streams of lattice points}
+
+In Hydroflow, a _lattice_\footnote{as in lattice order, from order
+theory} is a special class of datastructure equipped with a partial
+order ⊑ and _join_q (written ⊔) and _meet_ (written ⊓) functions.
+Both join and meet must be associative, commutative, and idempotent;
+additionally, join must inflate elements along ⊑, while meet must
+deflate them. That is: $∀ a,b: a ⊑ a ⊔ b ∧ b ⊑ a ⊔ b$ (for join) and
+$∀ a,b: a ⊓ b ⊑ a ∧ a ⊓ b ⊑ b$ (for meet).  Rather than operate on
+streams of arbitrary data types, hydroflow programs operate on streams
+of _Lattice Points_, or elements of lattices, drawn from any number of
+user-defined Lattices.
+
+Our choice to use lattices (and lattice points) provides key
+information to the hydroflow compiler when determining optimizations;
+which, of course, it does not take advantage of in any way whatsoever,
+so I guess this section needs to tie into the story about informing a
+higher-level analysis about lattice stuff.
+
+While our choice to use lattices alone might seem more restrictive
+than traditional dataflow engines, the reality is in fact the
+opposite; one can always build a lattice from an arbitrary datatype.
+Rather, our choice to privilege streams of lattice points provides the
+opportunity to the programmer to communicate certain datastructure
+features to the hydroflow compiler; should those features be absent,
+one can always take the free lattice over a given datatype. This is
+achieved by declaring all elements of that datatype to be unordered by
+⊑, and introducing an arbitrary ⊤ and ⊥ elements to serve as the
+supremum and infinimum of the lattice respectively.  One can then
+define $a ⊔ b ≜ (a == b ? a : ⊤)$ and $a ⊓ b ≜ (a == b ? a : ⊥)$,
+satisfying the requirements for both ⊔ and ⊓ operators.
+
+\subsubsection{Properties of streams}
+
 \bibliographystyle{ACM-Reference-Format}

### 2022-11-02 notes/types.tex
new directory
the axiom transformations continue to prove somewhat confusing.

--- 2022-10-26 types.tex
@@
         C ⊢ p : d \\ C';d ▷ op[0] ⊢ p'}
-                {{C,C'} ⊢ [p]\op[p']}}
+                {{C,C'} ⊢ (p)\op(p')}}
 
@@
       \inferrule[inward]{C ⊢ p : d}
-                {{C} ⊢ [p]\op : d ▷ \op[0]}}
+                {{C} ⊢ (p)\op : d ▷ \op[0]}}
 
@@
         C;d ▷ \op[0] ⊢ p'}
-                {{C};d ⊢ \op[p']}}
+                {{C};d ⊢ \op(p')}}
 
@@
 
-{\inferrule[axoim-transform-abstract]{Γ,dₐ▷A : τₐ ⊢ d ⇝ d' \\ A~\fresh}{Γ ⊢ d[A/dₐ] ⇝ d'}}
+%{\inferrule[axiom-transform-option1]{\forall s. d(s) \subseteq d'(s)}{Γ ⊢ d ⇝ d'}}
 
-{\inferrule[axoim-transform-concretize]{τₐ <: τ}{Γ,dₐ▷A : τₐ ⊢ d ⇝ d["?s"~τ/dₐ]}}
+{\inferrule[axiom-transform]{d \subseteq d'}{ ⊢ d ⇝ d'}}
 
-{\inferrule[axoim-transform-transitive]{Γ ⊢ d ⇝ d' \\ Γ ⊢ d' ⇝ d''}{
-Γ ⊢ d ⇝ d''}}
+%{\inferrule[axiom-transform-option3]{pre(d) \rightarrow pre(d') 
+% \\ post(d') \rightarrow post(d)}{Γ ⊢ d ⇝ d'}}
 
-{\inferrule[axoim-transform-unfold]{}{Γ,dₐ▷A : τₐ  ⊢ d ⇝ d[A/dₐ]}}
+     {\inferrule[axiom-transform-abstract]{Γ,dₐ▷A : τₐ ⊢ d ⇝ d' \\ A~\fresh}{Γ ⊢ d[A/dₐ] ⇝ d'}}
 
-{\inferrule[axoim-transform-fold]{}{Γ,dₐ▷A : τₐ  ⊢ d[A/dₐ] ⇝ d}}
+{\inferrule[axiom-transform-concretize]{τₐ <: τ \\ Γ ⊢ dₐ : τₐ}{Γ ⊢ d ⇝ d["?s"~τ/dₐ]}}
 
-{\inferrule[axoim-transform-reflexive]{}{Γ ⊢ d ⇝ d}}
+{\inferrule[axiom-transform-unfold]{}{Γ,dₐ▷A : τₐ  ⊢ d ⇝ d[A/dₐ]}}\MPM{ideally we'd find a way to do this one item at a time, not all-at-once like this. Also, to both sides of the ⇝.}
+
+{\inferrule[axiom-transform-fold]{}{Γ,dₐ▷A : τₐ  ⊢ d[A/dₐ] ⇝ d}}\MPM{Same comment here.}
 
@@
 
-Worth wondering if we're going to be in trouble overloading the $[]$ syntax here; time will tell. 
 Types appear in figure \ref{fig:types}. 
